
###`r set` 

####Vaccine comparison   {.tabset .tabset-fade .tabset-pills}

```{r}

## get significant gsea-sets
cam_vac %>%
  dplyr::filter(FDR <= fdr_cut) %>%
  pull(geneset)->
  vac_sets

## get genes for this geneset
set_genes <- geneIds(btm_gtm)[[set]]
delta %>%
  filter(genenames %in% set_genes) %>%
   mutate(disease=plyr::mapvalues(malaria_status,
                                 from=c(0,1), 
                        to=c("No", "Yes"))) %>%
  data.table() ->
  genes

## DEG analysis ranked sig to not_sig
sig_vac %>%
  mutate(genenames=rownames(sig_vac)) %>%
  filter(genenames %in% set_genes) %>%
  mutate(logFC=signif(logFC, 3)) %>%
  mutate(P.value=signif(P.value, 3)) %>%
  mutate(FDR=signif(FDR, 3)) %>%
  arrange(P.value) %>%
  dplyr::select(genenames, everything()) %>%
  mutate(index=1:length(logFC)) ->
  deg

## check deg names
if(length(setdiff(deg$genenames, set_genes)) != 0) {
  stop("deg plot gone wrong")
}

```
  
  
#####Table of significant genes  
  
Pvalue cutoff <= `r p_cut`    
    
```{r}
deg %>%
  dplyr::select(-c(index)) %>%
  filter(P.value <= p_cut) ->
  sig_deg

if(nrow(sig_deg) > 0) {
   sig_deg %>%
   headerKable()
} else {
  cat("No significant genes found")
}
all_sig_vac_genes <- c(all_sig_vac_genes, sig_deg$genenames)
```
  
#####Plot of all genes  
  
```{r}

##x <- c(0.05, seq(0.1,1, 0.1))
##pval_quants <- unlist(lapply(x, function(x, dat=deg$P.value) {return(which.min(dat <= x))}))

sig_genes <- deg$genenames[deg$P.value <= p_cut]
nummie <- length(sig_genes)
if(nummie == 0) {
  num_ind <- 0
} else {
  num_ind <- 1:nummie
}

ggplot(deg, aes(x=-log10(P.value), y=logFC)) + geom_point() + 
  ##gghighlight(P.Value <0.011, color="red", label_key=genenames,
  ##            label_params=list(color="red"))
  geom_point(data=deg[num_ind,], colour="red") +
  ##annotate("text", x=1+15, y=deg$logFC[1], label=deg$genenames[1], 
  ##         color="red") + 
  ##annotate("text", x=2+25, y=deg$logFC[2], label=deg$genenames[2], 
  ##         color="red") + 
  geom_hline(yintercept=0, linetype="dashed", color="black") +
  xlab("-log10(P.value") + 
  geom_vline(xintercept=-log10(p_cut), linetype="dashed", color="blue", size=0.1) +
 ##annotate("text", x=sig_ind+25, y=1, label="Pvalue <= 0.05", 
  ##         color="blue", size=4) +
  ##geom_line(aes(x=index, y=-c(1-P.value)), color="blue") +
  ##scale_y_continuous(sec.axis = sec_axis(~.+1, name = "Pvalue")) +
  theme( axis.line.y.right = element_line(color = "blue"), 
       axis.ticks.y.right = element_line(color = "blue"),
       axis.title.y.right= element_text(color = "blue"),
       axis.text.y.right= element_text(color = "blue"))
```
  
#####Plot signifcant genes  
  
```{r}
## get data for all significant genes in geneset

if(nummie > 0) {

  genes %>%
    filter(genenames %in% sig_genes) %>%
  mutate(disease=plyr::mapvalues(malaria_status,
                                 from=c(0,1), 
                        to=c("No", "Yes"))) %>%
  ##mutate(relevel(disease, "Yes")) %>%
  mutate(cell=factor(paste(vaccine, disease, sep="_"),
                     levels=c("rtss_Yes",
                              "rtss_No",
                              "comparator_Yes",
                              "comparator_No"))) %>%
  data.table() ->
plot_dt

  ggplot(plot_dt, aes(x=cell, y=stim_d, color=cell)) +
    geom_boxplot() + facet_wrap(~genenames, scales="free") +
    theme(axis.text.x=element_blank())
}
```


```{r}

if(nummie > 0) {
  
    ## plot means of each cell
    plot_means <- genes[,(mean = mean(get(stim_d), na.rm=TRUE)), 
                          by=list(genenames, vaccine, disease)]
    plot_means$disease <- relevel(plot_means$disease, "No")
    ggplot(plot_means, aes(x=disease, y=V1, group=vaccine)) + 
      geom_line(aes(col=vaccine)) +
      facet_wrap(~genenames, scales="free") +
      ylab("log2 CPM")
  
} 

```

####Disease comparison  {.tabset .tabset-fade .tabset-pills}  
  

```{r}

## get significant gsea-sets
cam_dis %>%
  dplyr::filter(FDR <= fdr_cut) %>%
  pull(geneset)->
  dis_sets


## DEG analysis ranked sig to not_sig
sig_dis %>%
  mutate(genenames=rownames(sig_vac)) %>%
  filter(genenames %in% set_genes) %>%
  mutate(logFC=signif(logFC, 3)) %>%
  mutate(P.value=signif(P.value, 3)) %>%
  mutate(FDR=signif(FDR, 3)) %>%
  arrange(P.value) %>%
  dplyr::select(genenames, everything()) %>%
  mutate(index=1:length(logFC)) ->
  deg

## check deg names
if(length(setdiff(deg$genenames, set_genes)) != 0) {
  stop("deg plot gone wrong")
}

```
  
  
#####Table of significant genes  
  
Pvalue cutoff <= `r p_cut`    
    
```{r}
deg %>%
  dplyr::select(-c(index)) %>%
  filter(P.value <= p_cut) ->
  sig_deg

if(nrow(sig_deg) > 0) {
   sig_deg %>%
   headerKable()
} else {
  cat("No significant genes found")
}
all_sig_vac_genes <- c(all_sig_vac_genes, sig_deg$genenames)
```
  
#####Plot of all genes  
  
```{r}

##x <- c(0.05, seq(0.1,1, 0.1))
##pval_quants <- unlist(lapply(x, function(x, dat=deg$P.value) {return(which.min(dat <= x))}))

sig_genes <- deg$genenames[deg$P.value <= p_cut]
nummie <- length(sig_genes)
if(nummie == 0) {
  num_ind <- 0
} else {
  num_ind <- 1:nummie
}

ggplot(deg, aes(x=-log10(P.value), y=logFC)) + geom_point() + 
  ##gghighlight(P.Value <0.011, color="red", label_key=genenames,
  ##            label_params=list(color="red"))
  geom_point(data=deg[num_ind,], colour="red") +
  ##annotate("text", x=1+15, y=deg$logFC[1], label=deg$genenames[1], 
  ##         color="red") + 
  ##annotate("text", x=2+25, y=deg$logFC[2], label=deg$genenames[2], 
  ##         color="red") + 
  geom_hline(yintercept=0, linetype="dashed", color="black") +
  xlab("-log10(P.value") + 
  geom_vline(xintercept=-log10(p_cut), linetype="dashed", color="blue", size=0.1) +
 ##annotate("text", x=sig_ind+25, y=1, label="Pvalue <= 0.05", 
  ##         color="blue", size=4) +
  ##geom_line(aes(x=index, y=-c(1-P.value)), color="blue") +
  ##scale_y_continuous(sec.axis = sec_axis(~.+1, name = "Pvalue")) +
  theme( axis.line.y.right = element_line(color = "blue"), 
       axis.ticks.y.right = element_line(color = "blue"),
       axis.title.y.right= element_text(color = "blue"),
       axis.text.y.right= element_text(color = "blue"))
```
  
#####Plot signifcant genes  
  
```{r}
## get data for all significant genes in geneset

if(nummie > 0) {

  genes %>%
    filter(genenames %in% sig_genes) %>%
  mutate(disease=plyr::mapvalues(malaria_status,
                                 from=c(0,1), 
                        to=c("No", "Yes"))) %>%
  ##mutate(relevel(disease, "Yes")) %>%
  mutate(cell=factor(paste(vaccine, disease, sep="_"),
                     levels=c("rtss_Yes",
                              "rtss_No",
                              "comparator_Yes",
                              "comparator_No"))) %>%
  data.table() ->
plot_dt

  ggplot(plot_dt, aes(x=cell, y=stim_d, color=cell)) +
    geom_boxplot() + facet_wrap(~genenames, scales="free") +
    theme(axis.text.x=element_blank())
}
```


```{r}

if(nummie > 0) {
  
    ## plot means of each cell
    plot_means <- genes[,(mean = mean(get(stim_d), na.rm=TRUE)), 
                          by=list(genenames, vaccine, disease)]
    plot_means$disease <- relevel(plot_means$disease, "No")
    ggplot(plot_means, aes(x=disease, y=V1, group=vaccine)) + 
      geom_line(aes(col=vaccine)) +
      facet_wrap(~genenames, scales="free") +
      ylab("log2 CPM")
  
} 

```

