
###`r set` {.tabset .tabset-fade .tabset-pills}

```{r}
## output geneset camera results
##vc_gsea[vc_gsea$geneset==set,] %>%
##  headerKable

genes <- mal_voom[btm_ind[[set]],]

## check gene names
check <- length(intersect(rownames(genes), geneIds(btm_gtm)[[set]]))
if(check != nrow(genes)) {stop("gene plot gone wrong")}

## DEG analysis ranked sig to not_sig
vc_deg %>%
  filter(genenames %in% rownames(genes$E)) %>%
  dplyr::select(-c("t", "B")) %>%
  mutate(logFC=signif(logFC, 3)) %>%
  mutate(AveExpr=signif(AveExpr, 3)) %>%
  mutate(P.Value=signif(P.Value, 3)) %>%
  mutate(adj.P.Val=signif(adj.P.Val, 3)) %>%
  arrange(P.Value) %>%
  dplyr::select(genenames, everything()) %>%
  mutate(index=1:length(logFC)) ->
  deg

## check deg names
if(length(setdiff(deg$genenames, rownames(genes))) != 0) {
  stop("deg plot gone wrong")
}
```
  
####Table of significant genes  
  
Pvalue cutoff <= `r p_cut`    
    
```{r}
deg %>%
  dplyr::select(-c(index)) %>%
  filter(P.Value <= p_cut) ->
  sig_deg

sig_deg %>%
headerKable()

all_sig_genes <- c(all_sig_genes, sig_deg$genenames)
```
  
####Plot of all genes  
  
```{r}

x <- c(0.05, seq(0.1,1, 0.1))
pval_quants <- unlist(lapply(x, function(x, dat=deg$P.Value) {return(which.min(dat <= x))}))

nummie <- sum(deg$P.Value <= p_cut)

ggplot(deg, aes(x=index, y=logFC)) + geom_point() + 
  ##gghighlight(P.Value <0.011, color="red", label_key=genenames,
  ##            label_params=list(color="red"))
  geom_point(data=deg[1:nummie,], colour="red") +
  ##annotate("text", x=1+15, y=deg$logFC[1], label=deg$genenames[1], 
  ##         color="red") + 
  ##annotate("text", x=2+25, y=deg$logFC[2], label=deg$genenames[2], 
  ##         color="red") + 
  geom_hline(yintercept=0, linetype="dashed", color="black") +
  xlab("gene") + 
  geom_vline(xintercept=pval_quants, linetype="dashed", color="blue", size=0.1) +
 ##annotate("text", x=sig_ind+25, y=1, label="Pvalue <= 0.05", 
  ##         color="blue", size=4) +
  geom_line(aes(x=index, y=-c(1-P.Value)), color="blue") +
  scale_y_continuous(sec.axis = sec_axis(~.+1, name = "Pvalue")) +
  theme( axis.line.y.right = element_line(color = "blue"), 
       axis.ticks.y.right = element_line(color = "blue"),
       axis.title.y.right= element_text(color = "blue"),
       axis.text.y.right= element_text(color = "blue"))
```
  
####Plot signifcant genes  
  
```{r}
## get data for all significant genes in geneset
deg %>%
  filter(P.Value <= p_cut) %$%
  genenames ->
sig_genes 

data.table::melt(genes$E[sig_genes,]) %>%
  dplyr::left_join(genes$targets[, c("col_id", "vaccine", "malaria_status")], by=c("Var2"="col_id")) %>%
  mutate(disease=plyr::mapvalues(malaria_status,
                                 from=c(0,1), 
                        to=c("No", "Yes"))) %>%
  ##mutate(relevel(disease, "Yes")) %>%
  mutate(cell=factor(paste(vaccine, disease, sep="_"),
                     levels=c("rtss_Yes",
                              "rtss_No",
                              "comparator_Yes",
                              "comparator_No"))) %>%
  data.table() ->
plot_dt

ggplot(plot_dt, aes(x=cell, y=value, color=cell)) +
  geom_boxplot() + facet_wrap(~Var1, scales="free") +
  theme(axis.text.x=element_blank())
```


```{r}

## plot means of each cell
plot_means <- plot_dt[,(mean = mean(value)), 
                      by=list(Var1, vaccine, malaria_status, disease)]
plot_means$disease <- relevel(plot_means$disease, "No")
ggplot(plot_means, aes(x=disease, y=V1, group=vaccine)) + 
  geom_line(aes(col=vaccine)) +
  facet_wrap(~Var1, scales="free") +
  ylab("log2 CPM")
  
##x <- plot_means[Var1=="SIGLEC9"]

```
