

```{r load-packages}
library(RNASeqUtilities)
library(Biobase)
library(GSEABase)
library(DT)
library(limma)
library(tidyverse)
library(magrittr)
library(here)
library(heatmap3)

source(here("code", "mal_utilities.R"))
```

```{r load-data}
library(mal067data)
data(mal067_eset)
meta_dt <- as_tibble(pData(mal067_eset))
```


```{r set-parameters}
SHOW_SAMPLE_SIZE <- TRUE

## parameters set by parent: stim, time

fdr_cut <- 0.2
p_cut <- 0.01
min_gene_set <- 5
form_interact_vac_both <- "~plate + total_reads + age + vaccine"
form_interact_dis_both <- "~plate + total_reads + age + case"
form_interact_vac <- "~plate + total_reads + vaccine"
form_interact_dis <- "~plate + total_reads + case"
##form_interact_dis <- "~plate + total_reads + age_weeks + malaria_status"

coef_dis <- "casecase"
##coef_dis <- "malaria_status1"

## store number of significant gene-sets for each comparison
summary_tab <- NULL

## if time is M0 set conditional to FALSE as there are no young subjects in M0
is_m0 <- time != "M0"
```

```{r genesets}
## set up GSEA analysis
btm_gtm <- getGmt(here("data/BTM_for_GSEA_20131008.gmt"))
btm_ind <- ids2indices(geneIds(btm_gtm), rownames(mal067_eset))
btm_ind <- btm_ind[sapply(btm_ind, length) > min_gene_set]
```

# GSEA `r toupper(stim)`: vaccine and disease main effects
  
**vaccine linear model:** `r form_interact_vac`  
**vaccine filtering   :**  `r time`
  
```{r vaccine_data}
################### vaccine and disease  analysis ########################

## vaccine: select subset of expressionSet
meta_dt %>% 
  filter(stimulation == stim) %>% 
  filter(visit == time) %$%
  col_id ->
  sample_vac


## generate subset data and drop extra stimulation levels
mal_vac <- mal067_eset[, sample_vac] 
mal_vac$stimulation <- fct_drop(mal_vac$stimulation)
```

**disease linear model:** `r form_interact_dis`  
**disease filtering   :** `r time`, case/control, rtss  

```{r disease_data}
## disease: select subset of expressionSet
meta_dt %>% 
  filter(stimulation == stim) %>% 
  filter(visit == time) %>%
  filter(case != "neither") %>%
  filter(vaccine == "rtss") %$%
  col_id ->
  sample_vac

## generate subset data and drop extra stimulation/case levels
mal_dis <- mal067_eset[, sample_vac] 
mal_dis$stimulation <- fct_drop(mal_dis$stimulation)
mal_dis$case <- fct_drop(mal_dis$case)

```

##both age groups (young and old) {.tabset .tabset-fade .tabset-pills}

```{r sample_size, eval=SHOW_SAMPLE_SIZE}

## get sample sizes
countSampleSizes(pData(mal_vac), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,3,2), 
            title="vaccine sample size")

## get sample sizes
countSampleSizes(pData(mal_dis), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,2,2), 
            title="disease sample size")

```
  
###vaccine  

```{r vac_analysis}
################## vaccine analysis ########################

cam_vac <- runGSEA(mal_vac, 
                   form_interact_vac_both,
                   coef="vaccinertss")
numSig <- sum(cam_vac$FDR <= fdr_cut)

suffix <- paste(stim, time, sep="_")
write_csv(cam_vac, paste0(here("output/"), "vaccine_both_", suffix, ".csv"))
```

###disease  

```{r dis_analysis, results="asis"}

################## disease analysis #######################

cam_dis <- runGSEA(mal_dis, 
                   form_interact_dis_both,
                   ##coef="casecase")
                   coef=coef_dis)
summary_tab <- rbind(c(numSig, sum(cam_dis$FDR <= fdr_cut)))


write_csv(cam_dis, paste0(here("output/"), "disease_both_", suffix, ".csv"))
```

###disease: significant vaccine gene-sets  
  
```{r dis_vac}

if(numSig == 0) {
  cat("No significant gene-sets for vaccine comparison")
} else {
  cam_vac %>%
    filter(FDR <= fdr_cut) %$%
    geneset ->
    vac_sigs
  
  cam_dis %>%
    filter(geneset %in% vac_sigs) %>%
    mutate(FDR = p.adjust(PValue, method="BH")) ->
    dis_out
  
  dis_out %>% 
   mutate(PValue=signif(PValue, 3)) %>%
    mutate(FDR=signif(FDR, 3)) %>%
    headerKable()
}

write_csv(dis_out, paste0(here("output/"), "disease_ds_both_", suffix, ".csv"))
```

###Heatmap  
  
```{r}

## combine allsignificant gene sets
cam_vac %>%
  filter(FDR <= fdr_cut) %>%
  filter(!grepl("TBA", geneset)) %>%
  dplyr::left_join(cam_dis[,c("geneset", "Direction", "PValue", "FDR")],
                   by="geneset") %>%
  plyr::rename(c("Direction.x"="Direction_vac")) %>%
  plyr::rename(c("PValue.x"="PValue_vac")) %>%
  plyr::rename(c("FDR.x"="FDR_vac")) %>%
  plyr::rename(c("Direction.y"="Direction_dis")) %>%
  plyr::rename(c("PValue.y"="PValue_dis")) %>%
  plyr::rename(c("FDR.y"="FDR_dis")) %>%
  mutate(sign_vac=as.numeric(plyr::mapvalues(Direction_vac, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(log10_Pval_vac=sign_vac*-log10(PValue_vac)) %>%
  mutate(sign_dis=as.numeric(plyr::mapvalues(Direction_dis, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
    mutate(log10_Pval_dis=sign_dis*-log10(PValue_dis)) ->
  outie

outie %>%
  filter(FDR_vac <= fdr_cut | FDR_dis <= fdr_cut) %>%
  filter(!grepl("TBA", geneset)) %$%
  geneset ->
  all_sigs

outie %>%
  ##filter(geneset %in% all_sigs) %>%
  dplyr::arrange(log10_Pval_vac) %>%
  dplyr::select(c(geneset, log10_Pval_vac, log10_Pval_dis)) %>%
  column_to_rownames("geneset") %>%
  plyr::rename(c(log10_Pval_vac="vaccine",
                 log10_Pval_dis="disease")) %>%
  as.matrix() ->
  heat_mat

heatmap3(heat_mat, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, 
         cexRow=0.50, cexCol=1.25, showRowDendro=FALSE, margins=c(5,8),
         col=colorRampPalette(c("navy", "white", "firebrick3"))(1024))

```

##young age group {.tabset .tabset-fade .tabset-pills}

```{r filter_young}

pData(mal_vac) %>%
  filter(age=="young") %$%
  col_id ->
  sample_vac_young

mal_vac_young <- mal_vac[, sample_vac_young] 

pData(mal_dis) %>%
  filter(age=="young") %$%
  col_id ->
  sample_dis_young

mal_dis_young <- mal_dis[, sample_dis_young] 

```


```{r sample_size_young, eval=SHOW_SAMPLE_SIZE}

## get sample sizes
countSampleSizes(pData(mal_vac_young), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,3,2), 
            title="vaccine sample size")

## get sample sizes
countSampleSizes(pData(mal_dis_young), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,2,2), 
            title="disease sample size")

```
  
###vaccine  

```{r vac_analysis_young, eval=is_m0}
################## vaccine analysis ############################

cam_vac_young <- runGSEA(mal_vac_young, 
                   form_interact_vac,
                   coef="vaccinertss")
numSig <- sum(cam_vac_young$FDR <= fdr_cut)

```

###disease  

```{r dis_analysis_young, eval=is_m0}

################## disease analysis ############################

cam_dis_young <- runGSEA(mal_dis_young, 
                   form_interact_dis,
                   coef=coef_dis)
summary_tab <- rbind(summary_tab, c(numSig, sum(cam_dis_young$FDR <= fdr_cut)))

if(stim=="dmso") {
    write_csv(cam_dis_young, path=here("output", 
                             paste(stim, time, "young", "disease.csv", sep="_")))
}
```

###disease: significant vaccine gene-sets  
  
```{r dis_vac_young, eval=is_m0}

if(numSig == 0) {
  cat("No significant gene-sets for vaccine comparison")
} else {
  cam_vac_young %>%
    filter(FDR <= fdr_cut) %$%
    geneset ->
    vac_sigs
  
  cam_dis_young %>%
    filter(geneset %in% vac_sigs) %>%
    mutate(FDR = p.adjust(PValue, method="BH"),
           FDR = signif(FDR, 4),
           PValue = signif(PValue, 4)) %>%
    kable(align=rep("l", ncol(cam_dis_young))) %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                      full_width=FALSE, position="left") %>%
        scroll_box(height="300px")
  
}
```

```{r set_zero, eval=!is_m0}
summary_tab <- rbind(summary_tab, c(NA, NA))
```


##old age group {.tabset .tabset-fade .tabset-pills}

```{r filter_old}

pData(mal_vac) %>%
  filter(age=="old") %$%
  col_id ->
  sample_vac_old

mal_vac_old <- mal_vac[, sample_vac_old] 

pData(mal_dis) %>%
  filter(age=="old") %$%
  col_id ->
  sample_dis_old

mal_dis_old <- mal_dis[, sample_dis_old] 

```


```{r sample_size_old, eval=SHOW_SAMPLE_SIZE}

## get sample sizes
countSampleSizes(pData(mal_vac_old), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,3,2), 
            title="vaccine sample size")

## get sample sizes
countSampleSizes(pData(mal_dis_old), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,2,2), 
            title="disease sample size")

```
  
###vaccine  

```{r vac_analysis_old}
################## vaccine analysis ############################

cam_vac_old <- runGSEA(mal_vac_old, 
                   form_interact_vac,
                   coef="vaccinertss")
numSig <- sum(cam_vac_old$FDR <= fdr_cut)

```

###disease  

```{r dis_analysis_old}

################## disease analysis ############################

cam_dis_old <- runGSEA(mal_dis_old, 
                   form_interact_dis,
                   coef=coef_dis)
summary_tab <- rbind(summary_tab, c(numSig, sum(cam_dis_old$FDR <= fdr_cut)))

if(stim=="dmso") {
    write_csv(cam_dis_old, path=here("output", 
                             paste(stim, time, "old", "disease.csv", sep="_")))
}
```

###disease: significant vaccine gene-sets  
  
```{r dis_vac_old}

if(numSig == 0) {
  cat("No significant gene-sets for vaccine comparison")
} else {
  cam_vac_old %>%
    filter(FDR <= fdr_cut) %$%
    geneset ->
    vac_sigs
  
  cam_dis_old %>%
    filter(geneset %in% vac_sigs) %>%
    mutate(FDR = p.adjust(PValue, method="BH"),
           FDR = signif(FDR, 4),
           PValue = signif(PValue, 4)) %>%
    kable(align=rep("l", ncol(cam_dis_old))) %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                      full_width=FALSE, position="left") %>%
        scroll_box(height="600px")
  
}

```

#Summary

```{r summary}
dimnames(summary_tab) <- list(c("both", "young", "old"), c("vaccine", "disease"))
headerKable(summary_tab, title="Number of significant gene-sets")
```
