

```{r}
library(RNASeqUtilities)
library(Biobase)
library(GSEABase)
library(DT)
library(limma)
library(tidyverse)
library(magrittr)
library(here)
library(data.table)

source(here("code", "mal_utilities.R"))
```

```{r}
library(mal067data)
data(mal067_eset)
meta_dt <- as_tibble(pData(mal067_eset))
```


```{r}
SHOW_SAMPLE_SIZE <- TRUE

## parameters set by parent: stim, 

fdr_cut <- 0.2
p_cut <- 0.01
min_gene_set <- 5
form_interact_vac_both <- "~plate + total_reads + age + pid + vaccine*stimulation"
form_interact_dis_both <- "~plate + total_reads + age + pid + malaria_status*stimulation"

form_interact_vac <- "~plate + total_reads + age_weeks + pid + vaccine*stimulation"
form_interact_dis <- "~plate + total_reads + age_weeks + pid + case*stimulation"

## store number of significant gene-sets for each comparison
summary_tab <- NULL
```

```{r genesets}
## set up GSEA analysis
btm_gtm <- getGmt(here("data/BTM_for_GSEA_20131008.gmt"))
btm_ind <- ids2indices(geneIds(btm_gtm), rownames(mal067_eset))
btm_ind <- btm_ind[sapply(btm_ind, length) > min_gene_set]
```

# GSEA `r toupper(stim)`: vaccine/disease stimulation interaction  
  
**vaccine linear model:** `r form_interact_vac`  
  
**vaccine filtering   :**  `r time`
  
```{r vaccine_data}
################### vaccine/disease*stimulation analysis ########################

## vaccine: select subset of expressionSet
meta_dt %>% 
  filter(stimulation == "dmso" | stimulation == stim) %>% 
  filter(visit == time) %$%
  col_id ->
  sample_vac

## generate subset data and drop extra stimulation levels
mal_vac <- mal067_eset[, sample_vac] 
mal_vac$stimulation <- fct_drop(mal_vac$stimulation)
```

**disease linear model:** `r form_interact_dis`   
  
**disease filtering   :** `r time`, case/control, rtss  

```{r disease_data}
## disease: select subset of expressionSet
meta_dt %>% 
  filter(stimulation == "dmso" | stimulation == stim) %>% 
  filter(visit == time) %>%
  filter(case != "neither") %>%
  filter(vaccine == "rtss") %$%
  col_id ->
  sample_vac

## generate subset data and drop extra stimulation/case levels
mal_dis <- mal067_eset[, sample_vac] 
mal_dis$stimulation <- fct_drop(mal_dis$stimulation)
mal_dis$case <- fct_drop(mal_dis$case)

```

##both age groups (young and old) {.tabset .tabset-fade .tabset-pills}

```{r sample_size, eval=SHOW_SAMPLE_SIZE}

## get sample sizes
countSampleSizes(pData(mal_vac), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,3,2), 
            title="vaccine sample size")

## get sample sizes
countSampleSizes(pData(mal_dis), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,2,2), 
            title="disease sample size")

```
  
###vaccine  

```{r vac_analysis}
################## vaccine analysis ############################

coeff <- paste0("vaccinertss:stimulation", stim)

## select subset of results from 'topTable' results
colInd <- c("logFC", "P.Value", "adj.P.Val")

cam_vac <- runGSEA(mal_vac, 
                   form_interact_vac,
                   coef=coeff)
numSig <- num_vac <- sum(cam_vac$FDR <= fdr_cut)

## apply DEG to determine driver genes
design <- model.matrix(formula(form_interact_vac), pData(mal_vac))
voom_vac <- voom(mal_vac, design)

 fit1 <- lmFit(voom_vac, design)
 fit2 <- eBayes(fit1, trend=FALSE)
 tmp <- topTable(fit2, number=Inf, coef=coeff, sort="P")
 sig_vac <- signif(tmp[, colInd], 3)
 colnames(sig_vac) <- c("logFC",  "P.value", "FDR")
 
 suffix <- paste(time, stim, sep="_")
 write_csv(tmp,  paste0(here("output/manuscript/"), "DEG_interaction_vaccine__", suffix, ".csv"))

```

###vaccine: no malaria

```{r}
################## vaccine analysis ############################

coeff <- paste0("vaccinertss:stimulation", stim)

## select subset of results from 'topTable' results
colInd <- c("logFC", "P.Value", "adj.P.Val")

## select only subjects without clinical malaria
mal_vac_dis <- mal_vac[,mal_vac$malaria_status==0]

cam_vac_dis <- runGSEA(mal_vac_dis, 
                   form_interact_vac,
                   coef=coeff)
```

###vaccine: malaria

```{r}
################## vaccine analysis ############################

coeff <- paste0("vaccinertss:stimulation", stim)

## select subset of results from 'topTable' results
colInd <- c("logFC", "P.Value", "adj.P.Val")

## select only subjects without clinical malaria
mal_vac_nodis <- mal_vac[,mal_vac$malaria_status==1]

cam_vac_nodis <- runGSEA(mal_vac_nodis, 
                   form_interact_vac,
                   coef=coeff)
```

###disease  

```{r dis_analysis, results="asis"}

################## disease analysis ############################

coeff <-paste0("casecase:stimulation", stim)
cam_dis <- runGSEA(mal_dis, 
                   form_interact_dis,
                   coef=coeff)
num_dis <- sum(cam_dis$FDR <= fdr_cut)
summary_tab <- rbind(c(num_vac, num_dis))

## apply DEG to determine driver genes
design <- model.matrix(formula(form_interact_dis), pData(mal_dis))
voom_dis <- voom(mal_dis, design)

 fit1 <- lmFit(voom_dis, design)
 fit2 <- eBayes(fit1, trend=FALSE)
 tmp <- topTable(fit2, number=Inf, coef=coeff, sort="P")
 tmp$gene <- rownames(tmp)
 sig_dis <- signif(tmp[, colInd], 3)
 colnames(sig_dis) <- c("logFC",  "P.value", "FDR")
 
 suffix <- paste(time, stim, sep="_")
 write_csv(tmp,  paste0(here("output/manuscript/"), "DEG_interaction_disease__", suffix, ".csv"))

```

```{r write_output}

cam_vac %>%
  dplyr::left_join(cam_dis[,c("geneset", "Direction", "PValue", "FDR")],
                   by="geneset") %>%
  plyr::rename(c("Direction.x"="Direction_vac")) %>%
  plyr::rename(c("PValue.x"="PValue_vac")) %>%
  plyr::rename(c("FDR.x"="FDR_vac")) %>%
  plyr::rename(c("Direction.y"="Direction_dis")) %>%
  plyr::rename(c("PValue.y"="PValue_dis")) %>%
  plyr::rename(c("FDR.y"="FDR_dis")) %>%
  mutate(sign_vac=as.numeric(plyr::mapvalues(Direction_vac, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(log10_Pval_vac=sign_vac*-log10(PValue_vac)) %>%
  mutate(sign_dis=as.numeric(plyr::mapvalues(Direction_dis, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
    mutate(log10_Pval_dis=sign_dis*-log10(PValue_dis)) ->
  outie
  
file <- paste("GSEA", time, stim, "vac_dis", sep="_")
write_csv(outie, paste0(here("output/"), file, ".csv"))

```

###disease: significant vaccine gene-sets  
  
```{r dis_vac}

if(num_vac == 0) {
  cat("No significant gene-sets for vaccine comparison")
} else {
  cam_vac %>%
    filter(FDR <= fdr_cut) %$%
    geneset ->
    vac_sigs
  
  cam_dis %>%
    filter(geneset %in% vac_sigs) %>%
    mutate(FDR = p.adjust(PValue, method="BH")) %>%
    kable(align=rep("c", ncol(cam_vac))) %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                      full_width=FALSE, position="left") %>%
        scroll_box(height="300px")
  
}

```




##young age group {.tabset .tabset-fade .tabset-pills}

```{r filter_young}

pData(mal_vac) %>%
  filter(age=="young") %$%
  col_id ->
  sample_vac_young

mal_vac_young <- mal_vac[, sample_vac_young] 

pData(mal_dis) %>%
  filter(age=="young") %$%
  col_id ->
  sample_dis_young

mal_dis_young <- mal_dis[, sample_dis_young] 

```


```{r sample_size_young, eval=SHOW_SAMPLE_SIZE}

## get sample sizes
countSampleSizes(pData(mal_vac_young), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,3,2), 
            title="vaccine sample size")

## get sample sizes
countSampleSizes(pData(mal_dis_young), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,2,2), 
            title="disease sample size")

```
  
###vaccine  

```{r vac_analysis_young}
################## vaccine analysis ############################

cam_vac_young <- runGSEA(mal_vac_young, 
                   form_interact_vac,
                   coef=paste0("vaccinertss:stimulation", stim))
numSig <- sum(cam_vac_young$FDR <= fdr_cut)

```

###disease  

```{r dis_analysis_young}

################## disease analysis ############################

cam_dis_young <- runGSEA(mal_dis_young, 
                   form_interact_dis,
                   coef=paste0("casecase:stimulation", stim))
summary_tab <- rbind(summary_tab, c(numSig, sum(cam_dis_young$FDR <= fdr_cut)))
```

###disease: significant vaccine gene-sets  
  
```{r dis_vac_young}

if(numSig == 0) {
  cat("No significant gene-sets for vaccine comparison")
} else {
  cam_vac_young %>%
    filter(FDR <= fdr_cut) %$%
    geneset ->
    vac_sigs
  
  cam_dis_young %>%
    filter(geneset %in% vac_sigs) %>%
    mutate(FDR = p.adjust(PValue, method="BH"),
           FDR = signif(FDR, 4),
           PValue = signif(PValue, 4)) %>%
    kable(align=rep("l", ncol(cam_dis_young))) %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                      full_width=FALSE, position="left") %>%
        scroll_box(height="300px")
  
}

```


##old age group {.tabset .tabset-fade .tabset-pills}

```{r filter_old}

pData(mal_vac) %>%
  filter(age=="old") %$%
  col_id ->
  sample_vac_old

mal_vac_old <- mal_vac[, sample_vac_old] 

pData(mal_dis) %>%
  filter(age=="old") %$%
  col_id ->
  sample_dis_old

mal_dis_old <- mal_dis[, sample_dis_old] 

```


```{r sample_size_old, eval=SHOW_SAMPLE_SIZE}

## get sample sizes
countSampleSizes(pData(mal_vac_old), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,3,2), 
            title="vaccine sample size")

## get sample sizes
countSampleSizes(pData(mal_dis_old), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,2,2), 
            title="disease sample size")

```
  
###vaccine  

```{r vac_analysis_old}
################## vaccine analysis ############################

cam_vac_old <- runGSEA(mal_vac_old, 
                   form_interact_vac,
                   coef=paste0("vaccinertss:stimulation", stim))
numSig <- sum(cam_vac_old$FDR <= fdr_cut)

```

###disease  

```{r dis_analysis_old}

################## disease analysis ############################

cam_dis_old <- runGSEA(mal_dis_old, 
                   form_interact_dis,
                   coef=paste0("casecase:stimulation", stim))
summary_tab <- rbind(summary_tab, c(numSig, sum(cam_dis_old$FDR <= fdr_cut)))
```

###disease: significant vaccine gene-sets  
  
```{r dis_vac_old}

if(numSig == 0) {
  cat("No significant gene-sets for vaccine comparison")
} else {
  cam_vac_old %>%
    filter(FDR <= fdr_cut) %$%
    geneset ->
    vac_sigs
  
  cam_dis_old %>%
    filter(geneset %in% vac_sigs) %>%
    mutate(FDR = p.adjust(PValue, method="BH"),
           FDR = signif(FDR, 4),
           PValue = signif(PValue, 4)) %>%
    kable(align=rep("l", ncol(cam_dis_old))) %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                      full_width=FALSE, position="left") %>%
        scroll_box(height="300px")
  
}

```

#Summary

```{r summary}
dimnames(summary_tab) <- list(c("both", "young", "old"), c("vaccine", "disease"))
headerKable(summary_tab, title="Number of significant gene-sets")
```


#Graph of deltas (stim - dmso) for disease vaccine interaction  
  
  
```{r}
## read in child file for rendering
child1 <- here::here("analysis/children", "gsea_plot_delta_genes.Rmd")
childy2 <- read_file(child1)
```

  
```{r}

all_sig_vac_genes <- all_sig_dis_genes <- NULL

## reconstruct deltas and put in form of an ExpressionSet
stim_d <- paste0(stim, "_d")

deltas <- readRDS(paste0(here::here(), "/output/stim_deltas.rds"))

deltas %>%
  dplyr::filter(visit==time) %>%
  dplyr::select(c("genenames", "col_id", stim_d)) %>%
  dplyr::left_join(mal067_voom$targets[,c("col_id", "vaccine", "malaria_status")],
            by="col_id") ->
  delta

```
  


```{r}
set <- cam_vac$geneset[1]
all_sig_genes <- NULL

```


`r paste(knit_child(text = childy2), collapse = '\n')`
</br>
   

```{r}
set <- cam_vac$geneset[2]
all_sig_genes <- NULL

```


`r paste(knit_child(text = childy2), collapse = '\n')`
</br>
   
```{r}
set <- cam_dis$geneset[1]
all_sig_genes <- NULL

```


`r paste(knit_child(text = childy2), collapse = '\n')`
</br>
   

```{r}
set <- cam_dis$geneset[2]
all_sig_genes <- NULL

```


`r paste(knit_child(text = childy2), collapse = '\n')`
</br>
 
#DEG analysis {.tabset .tabset-fade .tabset-pills}  
  
##vaccine  
  
```{r}

 summ <- sum(sig_vac$FDR <= fdr_cut)
 sig_vac$genes <- rownames(sig_vac)

 if(summ >0) {
     sig_vac %>% 
     filter(FDR <= fdr_cut) %>%
     select(genes, everything()) %>%
        datatable(rownames=FALSE)
  } else {
     cat("No significant gene-sets found")
  }
```

##disease  
  
```{r}

 summ <- sum(sig_dis$FDR <= fdr_cut)
 sig_dis$genes <- rownames(sig_dis)


 if(summ >0) {
     sig_dis %>%
     select(genes, everything()) %>%
     filter(FDR <= fdr_cut) %>%
     datatable(rownames=TRUE)
  } else {
     cat("No significant gene-sets found")
  }
```

  
  
