

```{r}
library(RNASeqUtilities)
library(Biobase)
library(GSEABase)
library(DT)
library(limma)
library(tidyverse)
library(magrittr)
library(here)
library(data.table)
library(heatmap3)

source(here("code", "mal_utilities.R"))
```

```{r}
library(mal067data)
data(mal067_eset)
meta_dt <- as_tibble(pData(mal067_eset))
```


```{r}
DO_PDF <- FALSE
SHOW_SAMPLE_SIZE <- TRUE

## parameters set by parent: stim, 

fdr_cut <- 0.2
p_cut <- 0.01
min_gene_set <- 5
form_interact_vac_both <- "~plate + total_reads + age_weeks + pid + stimulation*vaccine"
form_interact_dis_both <- "~plate + total_reads + age_weeks + pid + stimulation*case"

## store number of significant gene-sets for each comparison
summary_tab <- NULL
```

```{r genesets}
## set up GSEA analysis
btm_gtm <- getGmt(here("data/BTM_for_GSEA_20131008.gmt"))
btm_ind <- ids2indices(geneIds(btm_gtm), rownames(mal067_eset))
btm_ind <- btm_ind[sapply(btm_ind, length) > min_gene_set]
```

# GSEA `r toupper(stim)`: vaccine/case stimulation interaction  
  
**vaccine linear model:** `r form_interact_vac_both`  
  
**vaccine filtering   :**  `r time`
  
```{r vaccine_data}
################### vaccine/disease*stimulation analysis ########################

## vaccine: select subset of expressionSet
meta_dt %>% 
  filter(stimulation == "dmso" | stimulation == stim) %>% 
  filter(case != "neither") %>%
  filter(visit == time) %$%
  col_id ->
  sample_vac

## generate subset data and drop extra stimulation levels
mal_vac <- mal067_eset[, sample_vac] 
mal_vac$case <- fct_drop(mal_vac$case)
mal_vac$stimulation <- fct_drop(mal_vac$stimulation)

coeff <- paste0("stimulation", stim, ":vaccinertss")

## select subset of results from 'topTable' results
colInd <- c("logFC", "P.Value", "adj.P.Val")

```

**disease linear model:** `r form_interact_dis_both`   
  
**disease filtering   :** `r time`, case/control, rtss  

```{r disease_data}
## disease: select subset of expressionSet
meta_dt %>% 
  filter(stimulation == "dmso" | stimulation == stim) %>% 
  filter(visit == time) %>%
  filter(case != "neither") %>%
  filter(vaccine == "rtss") %$%
  col_id ->
  sample_vac

## generate subset data and drop extra stimulation/case levels
mal_dis <- mal067_eset[, sample_vac] 
mal_dis$stimulation <- fct_drop(mal_dis$stimulation)
mal_dis$case <- fct_drop(mal_dis$case)

```

##both age groups (infant and children) {.tabset .tabset-fade .tabset-pills}

```{r sample_size, eval=SHOW_SAMPLE_SIZE}

## get sample sizes
countSampleSizes(pData(mal_vac), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,2,2), 
            title="vaccine sample size")

## get sample sizes
countSampleSizes(pData(mal_dis), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,2,2), 
            title="disease sample size")

```
  
###vaccine  

```{r vac_analysis}
################## vaccine analysis ############################

cam_vac <- runGSEA(mal_vac, 
                   form_interact_vac_both,
                   coef=coeff)
numSig <- num_vac <- sum(cam_vac$FDR <= fdr_cut)

## apply DEG to determine driver genes
design <- model.matrix(formula(form_interact_vac_both), pData(mal_vac))
voom_vac <- voom(mal_vac, design)

 fit1 <- lmFit(voom_vac, design)
 fit2 <- eBayes(fit1, trend=FALSE)
 tmp <- topTable(fit2, number=Inf, coef=coeff, sort="P")
 sig_vac <- signif(tmp[, colInd], 3)
 colnames(sig_vac) <- c("logFC",  "P.value", "FDR")
 
 suffix <- paste(time, stim, sep="_")
 write_csv(tmp,  paste0(here("output/manuscript/"), "DEG_interaction_vaccine__", suffix, ".csv"))

```

###vaccine: case

```{r}
################## vaccine analysis ############################

## select only subjects without clinical malaria
mal_vac_dis <- mal_vac[,mal_vac$case=="case"]

cam_vac_dis <- runGSEA(mal_vac_dis, 
                   form_interact_vac_both,
                   coef=coeff)
```

###vaccine: control

```{r}
################## vaccine analysis ############################

## select only subjects without clinical malaria
mal_vac_nodis <- mal_vac[,mal_vac$case=="control"]

cam_vac_nodis <- runGSEA(mal_vac_nodis, 
                   form_interact_vac_both,
                   coef=coeff)
```

###disease: rts,s  

```{r dis_analysis, results="asis"}

################## disease analysis ############################

coeff <-paste0("stimulation", stim, ":casecase")
cam_dis <- runGSEA(mal_dis, 
                   form_interact_dis_both,
                   coef=coeff)
num_dis <- sum(cam_dis$FDR <= fdr_cut)
summary_tab <- rbind(c(num_vac, num_dis))

## apply DEG to determine driver genes
design <- model.matrix(formula(form_interact_dis_both), pData(mal_dis))
voom_dis <- voom(mal_dis, design)

 fit1 <- lmFit(voom_dis, design)
 fit2 <- eBayes(fit1, trend=FALSE)
 tmp <- topTable(fit2, number=Inf, coef=coeff, sort="P")
 tmp$gene <- rownames(tmp)
 sig_dis <- signif(tmp[, colInd], 3)
 colnames(sig_dis) <- c("logFC",  "P.value", "FDR")
 
 suffix <- paste(time, stim, sep="_")
 write_csv(tmp,  paste0(here("output/manuscript/"), "DEG_interaction_disease_rtss_", suffix, ".csv"))

```

###disease: comparator  

```{r disease_comp}
## disease: select subset of expressionSet
meta_dt %>% 
  filter(stimulation == "dmso" | stimulation == stim) %>% 
  filter(visit == time) %>%
  filter(case != "neither") %>%
  filter(vaccine == "comparator") %$%
  col_id ->
  sample_vac

## generate subset data and drop extra stimulation/case levels
mal_dis_comp <- mal067_eset[, sample_vac] 
mal_dis_comp$stimulation <- fct_drop(mal_dis_comp$stimulation)
mal_dis_comp$case <- fct_drop(mal_dis_comp$case)

################## disease analysis ############################

cam_dis_comp <- runGSEA(mal_dis_comp, 
                   form_interact_dis_both,
                   coef=coeff)
num_dis_comp <- sum(cam_dis_comp$FDR <= fdr_cut)
summary_tab <- rbind(c(num_vac, num_dis_comp))

## apply DEG to determine driver genes
design <- model.matrix(formula(form_interact_dis_both), pData(mal_dis_comp))
voom_dis <- voom(mal_dis_comp, design)

 fit1 <- lmFit(voom_dis, design)
 fit2 <- eBayes(fit1, trend=FALSE)
 tmp <- topTable(fit2, number=Inf, coef=coeff, sort="P")
 tmp$gene <- rownames(tmp)
 sig_dis <- signif(tmp[, colInd], 3)
 colnames(sig_dis) <- c("logFC",  "P.value", "FDR")
 
 suffix <- paste(time, stim, sep="_")
 write_csv(tmp,  paste0(here("output/manuscript/"),
                        "DEG_interaction_disease_comparator_", suffix, ".csv"))


```


```{r write_output}

cam_vac %>%
  dplyr::left_join(cam_dis[,c("geneset", "Direction", "PValue", "FDR")],
                   by="geneset") %>%
  plyr::rename(c("Direction.x"="Direction_vac")) %>%
  plyr::rename(c("PValue.x"="PValue_vac")) %>%
  plyr::rename(c("FDR.x"="FDR_vac")) %>%
  plyr::rename(c("Direction.y"="Direction_dis")) %>%
  plyr::rename(c("PValue.y"="PValue_dis")) %>%
  plyr::rename(c("FDR.y"="FDR_dis")) %>%
  mutate(sign_vac=as.numeric(plyr::mapvalues(Direction_vac, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(log10_Pval_vac=sign_vac*-log10(PValue_vac)) %>%
  mutate(sign_dis=as.numeric(plyr::mapvalues(Direction_dis, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
    mutate(log10_Pval_dis=sign_dis*-log10(PValue_dis)) ->
  outie
  
file <- paste("GSEA", time, stim, "vac_dis_case", sep="_")
write_csv(outie, paste0(here("output/manuscript/"), file, ".csv"))

```

###disease: rts,s, down-selected vaccine gene-sets  
  
```{r dis_vac_rtss}

if(num_vac == 0) {
  cat("No significant gene-sets for vaccine comparison")
} else {
  cam_vac %>%
    filter(FDR <= fdr_cut) %$%
    geneset ->
    vac_sigs
  
  cam_dis %>%
    filter(geneset %in% vac_sigs) %>%
    mutate(FDR = p.adjust(PValue, method="BH")) %>%
    kable(align=rep("c", ncol(cam_vac))) %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                      full_width=FALSE, position="left") %>%
        scroll_box(height="300px")
  
}

```

###disease: comparator, down-selected vaccine gene-sets  
  
```{r dis_vac_comp}

if(num_vac == 0) {
  cat("No significant gene-sets for vaccine comparison")
} else {
  cam_vac %>%
    filter(FDR <= fdr_cut) %$%
    geneset ->
    vac_sigs
  
  cam_dis_comp %>%
    filter(geneset %in% vac_sigs) %>%
    mutate(FDR = p.adjust(PValue, method="BH")) %>%
    kable(align=rep("c", ncol(cam_vac))) %>%
        kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                      full_width=FALSE, position="left") %>%
        scroll_box(height="300px")
  
}

```

```{r dis_output}

cam_dis %>%
  plyr::rename(c("Direction"="Direction_rtss")) %>%
  plyr::rename(c("PValue"="PValue_rtss")) %>%
  plyr::rename(c("FDR"="FDR_rtss")) %>%
  dplyr::select(-NGenes) %>%
  mutate(sign_vac_rtss=as.numeric(plyr::mapvalues(Direction_rtss, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(log10_Pval_rtss=sign_vac_rtss*-log10(PValue_rtss)) ->
    dis_rtss

cam_dis_comp %>%
  plyr::rename(c("Direction"="Direction_comp")) %>%
  plyr::rename(c("PValue"="PValue_comp")) %>%
  plyr::rename(c("FDR"="FDR_comp")) %>%
  dplyr::select(-NGenes) %>%
  mutate(sign_vac_comp=as.numeric(plyr::mapvalues(Direction_comp, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(log10_Pval_comp=sign_vac_comp*-log10(PValue_comp)) ->
    dis_comp

dis_rtss %>%
  dplyr::left_join(dis_comp, by="geneset") ->
  dis_out

file <- paste("GSEA", time, stim, "dis_case_by_vac", sep="_")
write_csv(dis_out, paste0(here("output/manuscript/"), file, ".csv"))

```

##Plots  
  
###heatmap of vaccine by disease (no down selection)
  
```{r} 

## read in BTM module annotations
btm_anno <- read_csv(here("data/btm_annotation_table_LNC.csv"))
btm_anno$full_title <- paste0(btm_anno$`Module title`, " (", btm_anno$ID, ")")

## get union of all significant 
all_sigs <- sort(Reduce(union, list(cam_vac$geneset[cam_vac$FDR <= fdr_cut],
                   cam_vac_dis$geneset[cam_vac_dis$FDR <= fdr_cut],
                   cam_vac_nodis$geneset[cam_vac_nodis$FDR <= fdr_cut],
                   cam_dis$geneset[cam_dis$FDR <= fdr_cut],
                   cam_dis_comp$geneset[cam_dis_comp$FDR <= fdr_cut])))

## don't make plots if there are no significant gene-sets
CONTINUE <- !is_empty(all_sigs)

```


```{r heat1, eval=CONTINUE}

## join genesets with their annotations
all_sigs <- dplyr::left_join(data.frame(full_title=all_sigs), 
                      btm_anno[,c("annotation", "full_title")],
                      by="full_title")

## extract -log10 signed pvalues for all genesets in all_sigs
format_gs <- function(mat, sigs=all_sigs, fdr=fdr_cut) {
  
  mat %>%
    arrange(geneset) %>%
    mutate(PValue=ifelse(FDR <= fdr, PValue, 1)) %>%
    mutate(FDR_sign=as.numeric(plyr::mapvalues(Direction, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
    mutate(pval_log10=-log10(PValue)*FDR_sign) %>%
    filter(geneset %in% sigs$full_title) ->
    tmp
  return(tmp)
}

sig_df <- data.frame(vaccine_all=format_gs(cam_vac)$pval_log10,
                     vaccine_case=format_gs(cam_vac_dis)$pval_log10,
                     vaccine_control=format_gs(cam_vac_nodis)$pval_log10,
                     disease_rtss=format_gs(cam_dis)$pval_log10,
                     disease_comp=format_gs(cam_dis_comp)$pval_log10)
sig_df$anno <- all_sigs$annotation
sig_df$geneset <- all_sigs$full_title

sig_df %>%
  filter(anno != "TBA") %>%
  arrange(anno, vaccine_all) %>%
  column_to_rownames("geneset") ->
  heat_mat

## colors for annotation
annos <- unique(heat_mat$anno)
anno_col <- rainbow(length(annos))
anno_cols <- anno_col[as.factor(heat_mat$anno)]

heat_mat$anno <- NULL

## only plot if there are significant results
  ##if(DO_PDF) {pdf(file=here("output/figures/manuscript/AMA1_heat_all_dis.pdf"))}
  
  heatmap3(heat_mat, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, 
           cexRow=0.75, cexCol=1.25, showRowDendro=FALSE, margins=c(2,20),
           col=colorRampPalette(c("navy", "white", "firebrick3"))(1024),
           RowSideColors=anno_cols, RowSideLabs="", na.rm=FALSE)
   legend(x="bottomleft", inset=-0.1, legend=annos, fill=anno_col, bty="n", cex=0.7)
 
##if(DO_PDF) {dev.off()}
 

```

###heatmap of vaccine and disease (with down selection)
  
```{r test2} 

## get list of down-selected genesets from vaccine comparison (both groups)
all_sigs <- unique(cam_vac$geneset[cam_vac$FDR <= fdr_cut])

## don't make plots if there are no significant gene-sets
CONTINUE <- !is_empty(all_sigs)
```

```{r heat2, eval=CONTINUE}

## join genesets with their annotations
all_sigs <- dplyr::left_join(data.frame(full_title=all_sigs), 
                      btm_anno[,c("annotation", "full_title")],
                      by="full_title")

## extract -log10 signed pvalues for all genesets in all_sigs
format_gs <- function(mat, sigs=all_sigs, fdr=fdr_cut) {
  
  mat %>%
    arrange(geneset) %>%
    mutate(PValue=ifelse(FDR <= fdr, PValue, 1)) %>%
    mutate(FDR_sign=as.numeric(plyr::mapvalues(Direction, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
    mutate(pval_log10=-log10(PValue)*FDR_sign) %>%
    filter(geneset %in% sigs$full_title) ->
    tmp
  return(tmp)
}

sig_df <- data.frame(vaccine_all=format_gs(cam_vac)$pval_log10,
                     disease_rtss=format_gs(cam_dis)$pval_log10,
                     disease_comp=format_gs(cam_dis_comp)$pval_log10)
sig_df$anno <- all_sigs$annotation
sig_df$geneset <- all_sigs$full_title

sig_df %>%
  filter(anno != "TBA") %>%
  arrange(anno, vaccine_all) %>%
  column_to_rownames("geneset") ->
  heat_mat

## colors for annotation
annos <- unique(heat_mat$anno)
anno_col <- rainbow(length(annos))
anno_cols <- anno_col[as.factor(heat_mat$anno)]

heat_mat$anno <- NULL

  ##if(DO_PDF) {pdf(file=here("output/figures/manuscript/AMA1_heat_all_dis.pdf"))}
  
  heatmap3(heat_mat, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, 
           cexRow=0.75, cexCol=1.25, showRowDendro=FALSE, margins=c(2,20),
           col=colorRampPalette(c("navy", "white", "firebrick3"))(1024),
           RowSideColors=anno_cols, RowSideLabs="", na.rm=FALSE)
   legend(x="bottomleft", inset=-0.1, legend=annos, fill=anno_col, bty="n", cex=0.7)
   
  ##if(DO_PDF) {dev.off()}


```






