---
title: "GSEA analysis stimulations"
author: "Carl Murie"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---


```{r load-packages, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)

library(Biobase)
library(tidyverse)
library(data.table)
library(here)
library(RNASeqUtilities)
library(DT)
library(magrittr)
library(readr)
library(venn)
library(heatmap3)
```

#Comparison of GSEA results for the three stimulation interactions (stim-DMSO)  
  
For each stimulation a test was applied to the vaccine effect (rts,s and comparator) and disease effect (protected vs not protected).
  
Heatmaps show all significant gene sets for both disease and vaccine comparisons together for all three stimulations.   
  
Scale is signed log10 PValue where blue is up-regulated and red is down-regulated.
  
##M3  {.tabset .tabset-fade .tabset-pills}
  

```{r load_data_M3}
ama  <- read_csv(here("output/GSEA_M3_ama1_vac_dis.csv"))
csp  <- read_csv(here("output/GSEA_M3_csp_vac_dis.csv"))
hbs  <- read_csv(here("output/GSEA_M3_hbs_vac_dis.csv"))
fdr_cut <- 0.2
```

```{r combine_M3}
ama %>%
  filter(FDR_vac <= fdr_cut | FDR_dis <= fdr_cut) %$%
  geneset ->
ama_sigs 

csp %>%
  filter(FDR_vac <= fdr_cut | FDR_dis <= fdr_cut) %$%
  geneset ->
csp_sigs 

hbs %>%
  filter(FDR_vac <= fdr_cut | FDR_dis <= fdr_cut) %$%
  geneset ->
hbs_sigs 

all_sigs <- unique(c(ama_sigs, csp_sigs, hbs_sigs))
```

###vaccine

```{r vac_m3}

## get log10Pval for vaccine effect for all stimulations
ama %>%
  filter(geneset %in% all_sigs) %>%
  dplyr::select(c(geneset, c("geneset", "log10_Pval_vac"))) %>%
  dplyr::left_join(csp[,c("geneset", "log10_Pval_vac")], by="geneset") %>%
  dplyr::left_join(hbs[,c("geneset", "log10_Pval_vac")], by="geneset") %>%
  plyr::rename(c("log10_Pval_vac.x"="ama1", 
                 "log10_Pval_vac.y"="csp",
                 "log10_Pval_vac"="hbs")) %>%
  column_to_rownames("geneset") %>%
  as.matrix() ->
  heat_mat

## get FDR for vaccine effect for all stimulations
ama %>%
  filter(geneset %in% all_sigs) %>%
  dplyr::select(c(geneset, c("geneset", "FDR_vac"))) %>%
  dplyr::left_join(csp[,c("geneset", "FDR_vac")], by="geneset") %>%
  dplyr::left_join(hbs[,c("geneset", "FDR_vac")], by="geneset") %>%
  plyr::rename(c("FDR_vac.x"="ama1", 
                 "FDR_vac.y"="csp",
                 "FDR_vac"="hbs")) %>%
  column_to_rownames("geneset") %>%
  as.data.frame() ->
  fdr_mat
  
  fdr_index <- which(fdr_mat <= fdr_cut, arr.ind=TRUE)
  fdr_index <- cbind(data.frame(fdr_index), "yellow", 5)
 

heatmap3(heat_mat, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, highlightCell=fdr_index,
         cexRow=1.5 ,showRowDendro=FALSE, 
         col=colorRampPalette(c("firebrick3", "white", "navy"))(1024))
```

###disease

```{r dis_m3}

## get log1Pval for vaccine effect for all stimulations
ama %>%
  filter(geneset %in% all_sigs) %>%
  dplyr::select(c(geneset, c("geneset", "log10_Pval_dis"))) %>%
  dplyr::left_join(csp[,c("geneset", "log10_Pval_dis")], by="geneset") %>%
  dplyr::left_join(hbs[,c("geneset", "log10_Pval_dis")], by="geneset") %>%
  plyr::rename(c("log10_Pval_dis.x"="ama1", 
                 "log10_Pval_dis.y"="csp",
                 "log10_Pval_dis"="hbs")) %>%
  column_to_rownames("geneset") %>%
  as.matrix() ->
  heat_mat

## get FDR for vaccine effect for all stimulations
ama %>%
  filter(geneset %in% all_sigs) %>%
  dplyr::select(c(geneset, c("geneset", "FDR_dis"))) %>%
  dplyr::left_join(csp[,c("geneset", "FDR_dis")], by="geneset") %>%
  dplyr::left_join(hbs[,c("geneset", "FDR_dis")], by="geneset") %>%
  plyr::rename(c("FDR_dis.x"="ama1", 
                 "FDR_dis.y"="csp",
                 "FDR_dis"="hbs")) %>%
  column_to_rownames("geneset") %>%
  as.data.frame() ->
  fdr_mat
  
  fdr_index <- which(fdr_mat <= fdr_cut, arr.ind=TRUE)
  fdr_index <- cbind(data.frame(fdr_index), "yellow", 5)
 


heatmap3(heat_mat, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, highlightCell=fdr_index,
         cexRow=1.5 ,showRowDendro=FALSE, 
         col=colorRampPalette(c("firebrick3", "white", "navy"))(1024))
```

