---
title: Combined Analysis
subtitle:  mal067, Vahey, and Kazmin
author: Carl Murie
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
  html_document:
    toc: true
    toc_float: true
    number_sections: true
---

```{r load-packages, echo=FALSE, warning=FALSE, message=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message=FALSE)

library(Biobase)
library(tidyverse)
library(data.table)
library(here)
library(RNASeqUtilities)
library(DT)
library(magrittr)
library(readr)
library(venn)
library(heatmap3)
```

#Comparison of three RTS,S data sets  

Three data sets will be compared:  
  
1. mal067 RTS,S RNASeq assay
2. Expression of Genes Associated with Immunoproteaseome Processing of Major Histocompabtibility Complex Peptides is Indicative of Protection with Adjuvanated RTS,S Malaria Vaccine, Vahey et al. 2010 (microrarray assay)
3. Systems analysis of protective immune responses to RTS,S malaria vaccination in humans, Kazmin et al. 2017 (microarray assay)  

**Subjects**  
  
1. Mal067: Subjects were children (6w -17m) drawn from Bagamoyo or Manhica, Africa  
  
2. Vahey et al.: Subjects were vaccine naive and adults drawn from Walter Reed Army Institute of Research  
  
3. Kazmin et al.: Subjects were vaccine naive and adults drawn from Walter Reed Army Institute of Research  
  
**GSEA analysis** was applied at two timepoints:  
  
1. Baseline - before vaccination
2. After complete vaccination treatment and challenge 
  
  + mal067: 3 months after baseline (~90 days) 
  + Vahey et al: ( 5 days after challenge)
  + Kazmin et al: 82 days after baseline (5 days after challenge)  
  
GSEA comparison was (NoProtection - Protection) or (case - control) 
  
Both age groups were combined for the mal067 analysis  
  
```{r load_data}
pul  <- read_csv(here("data/Pulendran_GSEA_disease.csv"))
vahey <- read_csv(here("data/Vahey_GSEA_disease.csv"))
mal_m0 <- read_csv(here("output/dmso_M0_both_disease.csv"))
mal_m3 <- read_csv(here("output/dmso_M3_both_disease.csv"))

fdr_cut <- 0.2
DO_VENN <- TRUE
```

FDR cutoff was <= `r fdr_cut`  
  
##Baseline  
  
```{r setup_m0}
pul %>%
  filter(visit=="D00") %>%
  dplyr::select(-visit) %>%
  mutate(assay="kazmin") ->
  pul_m0

vahey %>%
  filter(time=="T0") %>%
  dplyr::select(-time) %>%
  mutate(assay="vahey") ->
  vah_m0

mal_m0 %<>%
  mutate(assay="mal067")
```

```{r select_sigs_m0}
pul_m0 %>%
  filter(FDR <= fdr_cut) %$%
  geneset ->
pul_sigs

vah_m0 %>%
  filter(FDR <= fdr_cut) %$%
  geneset ->
vah_sigs

mal_m0 %>%
  filter(FDR <= fdr_cut) %$%
  geneset ->
  mal_sigs

all_sets <- rbind(mal_m0, vah_m0, pul_m0)
```

###Venn diagram of overlap  
  
```{r venn_m0}
venn(list(mal067=mal_sigs, vahey=vah_sigs, kazmin=pul_sigs),zcolor = "style")
```

###Heatmap of overlap  
  
```{r heat_M0}
## get union of all pairwise significant intersections
sig_union <- unique(c(intersect(mal_sigs, vah_sigs),
                      intersect(mal_sigs, pul_sigs),
                      intersect(vah_sigs, pul_sigs)))

## get union and calculate signed log10 FDR
all_sets %>%
  filter(geneset %in% sig_union) %>%
  mutate(sign = as.numeric(plyr::mapvalues(Direction, 
                                           from=c("Up", "Down"), 
                                           to=c(1, -1)))) %>%
  mutate(FDR_sign = -log10(FDR)*sign) ->
  all_sig

dcast(all_sig, geneset~assay, value.var="FDR_sign") %>%
 as_tibble() %>%
  column_to_rownames("geneset") %>%
  as.matrix() ->
  heat_mat

heatmap3(heat_mat, Rowv=TRUE, Colv=NA, scale="none", balanceColor=TRUE, 
         cexRow=0.65 ,showRowDendro=FALSE, 
         col=colorRampPalette(c("firebrick3", "white", "navy"))(1024))
```
  
###Table  
  
```{r data_m0}
all_sig %>%
  dplyr::select(-c("sign", "FDR_sign")) %>%
  datatable() %>%
  formatRound(columns=c("PValue", "FDR"), digit=3)
```
  
##After challenge  
  
```{r setup_m3}
pul %>%
  filter(visit=="D82") %>%
  dplyr::select(-visit) %>%
  mutate(assay="kazmin") ->
  pul_m3

vahey %>%
  filter(time=="T5") %>%
  dplyr::select(-time) %>%
  mutate(assay="vahey") ->
  vah_m3

mal_m3 %<>%
  mutate(assay="mal067")


```

```{r select_sigs}
pul_m3 %>%
  filter(FDR <= fdr_cut) %$%
  geneset ->
pul_sigs

vah_m3 %>%
  filter(FDR <= fdr_cut) %$%
  geneset ->
vah_sigs

mal_m3 %>%
  filter(FDR <= fdr_cut) %$%
  geneset ->
  mal_sigs

all_sets <- rbind(mal_m3, vah_m3, pul_m3)
```

###Venn diagram of overlap  
  
```{r venn_m3, out.width='75%'}
venn(list(mal067=mal_sigs, vahey=vah_sigs, kazmin=pul_sigs), zcolor = "style")

```

###Heatmap of overlap  
  
```{r heat_M3}
## get union of all pairwise significant intersections
sig_union <- unique(c(intersect(mal_sigs, vah_sigs),
                      intersect(mal_sigs, pul_sigs),
                      intersect(vah_sigs, pul_sigs)))

## get union and calculate signed log10 FDR
all_sets %>%
  filter(geneset %in% sig_union) %>%
  mutate(sign = as.numeric(plyr::mapvalues(Direction, 
                                           from=c("Up", "Down"), 
                                           to=c(1, -1)))) %>%
  mutate(FDR_sign = -log10(FDR)*sign) ->
  all_sig

dcast(all_sig, geneset~assay, value.var="FDR_sign") %>%
 as_tibble() %>%
  column_to_rownames("geneset") %>%
  as.matrix() ->
  heat_mat

##heat_mat[,1] <- -1*heat_mat[,1]

heatmap3(heat_mat, Rowv=TRUE, Colv=NA, scale="none", balanceColor=TRUE, 
         cexRow=0.65 ,showRowDendro=FALSE,
         col=colorRampPalette(c("firebrick3", "white", "navy"))(1024))
```
 
###Heatmap diagram of Vayey vs Pulendran  
  
```{r vp}
## get union of all pairwise significant intersections
sig_union <- unique(union(vah_sigs, pul_sigs))

## get union and calculate signed log10 FDR
all_sets %>%
  filter(geneset %in% sig_union) %>%
  filter(assay != "mal067") %>%
  mutate(sign = as.numeric(plyr::mapvalues(Direction, 
                                           from=c("Up", "Down"), 
                                           to=c(1, -1)))) %>%
  mutate(FDR_sign = -log10(FDR)*sign) ->
  all_sig

dcast(all_sig, geneset~assay, value.var="FDR_sign") %>%
 as_tibble() %>%
  column_to_rownames("geneset") %>%
  filter(!is.na(kazmin) & !is.na(vahey)) %>%
  as.matrix() ->
  heat_mat

```
  
**signs are normal**  
  
```{r}
heatmap3(heat_mat, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, 
         cexRow=0.65 ,showRowDendro=FALSE,
         col=colorRampPalette(c("firebrick3", "white", "navy"))(1024))
```

**Kazmin signs are reversed**

```{r}
heat_mat[,1] <- -1*heat_mat[,1]

heatmap3(heat_mat, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, 
         cexRow=0.65 ,showRowDendro=FALSE,
         col=colorRampPalette(c("firebrick3", "white", "navy"))(1024))
```

  
###Table  
  
```{r data_m3}
all_sig %>%
  dplyr::select(-c("sign", "FDR_sign")) %>%
  datatable() %>%
  formatRound(columns=c("PValue", "FDR"), digit=3)
```
 
