---
title: "MAL067: Experimental Design"
author: "Carl Murie and Raphael Gottardo"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
   BiocStyle::html_document:
    theme: lumen
    toc: true
    toc_float: true
    number_sections: true
---

```{r overall-knitr-options, echo=FALSE, results='hide'}
library(knitr)
opts_chunk$set(cache=FALSE, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results='asis')
```

```{r packages}
library(RNASeqUtilities)
library(Biobase)
library(tidyverse)
```


```{r load-data}
library(mal067data)
data(mal067_eset)
meta_dt <- as_tibble(pData(mal067_eset))

numberOfSubs <- length(unique(meta_dt$pid))
```

#Experimental Design

Bulk RNASeq RTS,S malaria vaccine study. 

1. Investigate mechanisms of the RTS,S vaccine-induced protection against malaria  
2. Identify a correlate of RTS,S vaccine-induced protection  

**Subjects**: `r numberOfSubs` subjects.

**Vaccine**: rtss and comparator

**Stimulation**: dmso, csp, hbs, ama1

**Location**: bagamoyo and manhica

**Age**: 6-12 weeks (young) and 5-17 months (old).

**Outcome**: control (no malaria) and case (malaria) which is a subset of all subjects that have an associated stratified variable.  
  
**Malaria status**: Whether subject contracted malaria in the following year  

**Visit**: M0 and M3 (initial visit and 3 months later)


#Sequencing

Reads were aligned using BWA Aln version 0.7.10. The transcriptome was downloaded from
UCSC RefSeq (Human 19 or Mouse 10). Mitochondrial genes and ERCC reference sequences were
added. Sequence alignment and quantification were run at Broad Technology Labs. The data was voom transformed and a linear model was applied for statistical testing.

Samples with total read counts less than 75000 were removed. Genes that did not have at least 20 samples (approximately 10% of the sample size) with read counts greater than 5 were removed. 

There were `r ncol(mal067_eset)` samples and `r nrow(mal067_eset)` genes remaining after filtering. 


#Sample Sizes

There is a significant confound between age and location. Bagamoyo had samples only from the old age group and Manhica had mostly samples from the young age group.

##By location  
  
```{r samples_site}
meta_dt %>%
  filter(site=="BAGAMOYO") %>%
  countSampleSizes("stimulation", c("age", "visit", "vaccine", "case")) %>%
  headerKable(labels=c("age", "visit", "vaccine", "disease"), levels=c(2,2,2,3),
              title="Bagamoyo")

meta_dt %>%
  filter(site=="MANHICA") %>%
  countSampleSizes("stimulation", c("age", "visit", "vaccine", "case")) %>%
  headerKable(labels=c("age", "visit", "vaccine", "disease"), levels=c(2,2,2,3),
              title="Manhica")

```
  
##By Plate  
  
```{r samples_plate}
meta_dt %>%
  filter(site=="BAGAMOYO") %>%
  countSampleSizes("plate", c("stimulation","age","visit","vaccine","case")) %>%
  headerKable(labels=c("stimulation", "age", "visit", "vaccine", "disease"),
              levels=c(4,2,2,2,3),
              title="Bagamoyo")

meta_dt %>%
  filter(site=="MANHICA") %>%
  countSampleSizes("plate", c("stimulation","age","visit","vaccine","case")) %>%
  headerKable(labels=c("stimulation", "age", "visit", "vaccine", "disease"),
              levels=c(4,2,2,2,3),
              title="Manhica")

```
  
#Protection

```{r protection}
subjects <- meta_dt[!duplicated(meta_dt$pid),]

countSampleSizes(subjects, "vaccine", "malaria_status") %>%
  dplyr::rename(Protected="X0", nonProtected="X1") %>%
  mutate(vaccine=c("comparator", "rts,s")) %>%
  select(vaccine, everything()) %>%
  mutate('protection (%)'=signif(100*(Protected/total), 3)) ->
           prot

## by age
subjects %>%
  filter(age=="young") %>%
countSampleSizes("vaccine", "malaria_status") %>%
  dplyr::rename(Protected="X0", nonProtected="X1") %>%
  mutate(vaccine=c("comparator", "rts,s")) %>%
  select(vaccine, everything()) %>%
  mutate('protection (%)'=signif(100*(Protected/total), 3)) ->
           prot_young

subjects %>%
  filter(age=="old") %>%
countSampleSizes("vaccine", "malaria_status") %>%
  dplyr::rename(Protected="X0", nonProtected="X1") %>%
  mutate(vaccine=c("comparator", "rts,s")) %>%
  select(vaccine, everything()) %>%
  mutate('protection (%)'=signif(100*(Protected/total), 3)) ->
           prot_old


subjects %>%
  filter(case != "neither") %>%
  countSampleSizes("vaccine", "malaria_status") %>%
  mutate(vaccine=c("comparator", "rts,s")) %>%
  select(vaccine, everything()) %>%
  dplyr::rename(Protected="X0", nonProtected="X1") %>%
  mutate('protection (%)'=signif(100*(Protected/total), 3)) ->
           prot_case

subjects %>%
  filter(case != "neither") %>%
  filter(age=="young") %>%
  countSampleSizes("vaccine", "malaria_status") %>%
  mutate(vaccine=c("comparator", "rts,s")) %>%
  select(vaccine, everything()) %>%
  dplyr::rename(Protected="X0", nonProtected="X1") %>%
  mutate('protection (%)'=signif(100*(Protected/total), 3)) ->
           prot_case_young

subjects %>%
  filter(case != "neither") %>%
  filter(age=="old") %>%
  countSampleSizes("vaccine", "malaria_status") %>%
  mutate(vaccine=c("comparator", "rts,s")) %>%
  select(vaccine, everything()) %>%
  dplyr::rename(Protected="X0", nonProtected="X1") %>%
  mutate('protection (%)'=signif(100*(Protected/total), 3)) ->
           prot_case_old


```

##All subjects  
  
###both age groups  
```{r}
prot %>%
  headerKable(title="both age groups")
```

###young age group    
```{r}
prot_young %>%
  headerKable(title="young age group")
```

###old age group  
```{r}
prot_old %>%
  headerKable(title="old age group")
```

##case/control subjects  
  
###both age groups
```{r}
prot_case %>%
  headerKable(title="both age groups")
```

###young age group
```{r}
prot_case_young %>%
  headerKable(title="young age group")
```

###old age group
```{r}
prot_case_old %>%
  headerKable(title="old age group")
```


