---
title: "Integrate antibody and RNASeq data"
author: "Carl Murie and Raphael Gottardo"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
   BiocStyle::html_document:
    theme: lumen
    toc: true
    toc_float: true
    number_sections: true
---

```{r overall-knitr-options, echo=FALSE, results='hide'}
library(knitr)
opts_chunk$set(cache=FALSE, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results='asis')
```

```{r packages}
library(RNASeqUtilities)
library(Biobase)
library(limma)
library(magrittr)
library(here)
library(data.table)
library(survival)   ## clogit
library(GSEABase)
library(tidyverse)
library(htmltools)
library(limma)

source(here("code", "mal_utilities.R"))
```

```{r load-data}
library(mal067data)
data(mal067_eset)
meta_dt <- as_tibble(pData(mal067_eset))
```


```{r gsea}
min_gene_set <- 5

## set up GSEA analysis
btm_gtm <- getGmt(here("data/BTM_for_GSEA_20131008.gmt"))
btm_ind <- ids2indices(geneIds(btm_gtm), rownames(mal067_eset))
btm_ind <- btm_ind[sapply(btm_ind, length) > min_gene_set]
```


```{r getmorevars}

## l_df_flow and l_df_rna
##
## 1. cytokines: stimulation (HBS, CSP),antigens (28), and log10 stim/DMSO ratios
## 2. IAVI: IgG (CSP - NANP), IgG (CSP - C-term)
## 3. CEVAC: IgG CSP-NANP titers, IgG HBsAg titers (both in log10 of the titers)
## 4. antibody Luminex: isotope (IgG, IgM), analyte (42 antigens), log10 of the
##                      median fluorecence intensity
load("data/m067_seattle_data.RData")

```

# Antibodies and RNASeq analysis  
  
Ddatabase of antibody Luminex contains 3 new variables: isotype (IgG and IgM), analyte (42 antigens) and log10_mfi_no_dil, as log10 of the median fluorecence intensity (mfi) not diluted. 
  
These variables will be applied to the DMSO samples.  
  
```{r cytokines} 

## format rnaseq meta data
meta_dt %>%
  filter(vaccine=="rtss") %>%
  filter(stimulation=="dmso") %>%
  mutate(stimulation=as.character(stimulation)) %>%
  mutate(visit=as.character(visit)) %>%
  dplyr::select(pid, visit, stimulation, col_id, case, malaria_status, match, plate) ->
  meta_cyto

## format antibody data
l_df_rna[[4]] %>%
  filter(vaccine=="rtss") %>%
  mutate(visit=toupper(visit))->
  df_rna

##x <- df_rna[df_rna$analyte=="alfagal",]
##countSampleSizes(x, "visit", c("isotype")) %>%
##  headerKable()

## join cytokine data with rnaseq meta data
dplyr::inner_join(df_rna, meta_cyto, by=c("pid", "visit")) %>%
dplyr::rename(log10_mfi=log10_mfi_no_dil) ->
  cyto

## check that total sample size is correct
total <- length(intersect(paste(meta_cyto$pid,meta_cyto$visit),
                          paste(df_rna$pid, df_rna$visit))) *
  length(unique(df_rna$analyte)) * 
  length(unique(df_rna$isotype))
if(total != nrow(cyto)) {stop("Sample size is not correct")}

```

## Sample size

###antibody database  
  
Each analyte has the following sample sizes  
  
```{r samplesize1}
##countSampleSizes(df_rna, "analyte", c("isotype", "visit", "vaccine")) %>%
##  headerKable()

tmp <- df_rna[df_rna$analyte=="alfagal",]
countSampleSizes(tmp, "isotype",  c("visit", "vaccine")) %>%
  headerKable()
```
 

# Disease prediction  

A clogit test of case-control vs mfi with gender and match as a stratified variable was applied to the antibody mfi.  

```{r cyto_logit}

## gather just case/control samples which the match stratified variable
cyto %>%
  filter(case != "neither") %>%
  mutate(case = fct_drop(case)) %>%                    ## remove "neither" level
  ## clogit needs case as a binary value (0,1)
  mutate(case_logit=as.numeric(plyr::mapvalues(as.character(case), 
                                               from=c("control", "case"),
                                               to=c(0,1)))) %>%
  as.data.table() ->
  cyto_logit
```

##Sample size  
  
No M0 samples were present after filtering for case/control so only the M3 time point will be analysed.
  
```{r sample3}
countSampleSizes(cyto_logit, "analyte", c("visit", "isotype", "case")) %>%
  headerKable()
```

```{r output}
 pvals <- cyto_logit[,summary(clogit(case_logit~sex+log10_mfi+strata(match)))$coefficient["log10_mfi", 5], by=.(analyte, isotype)]
 
  ## plot boxplots of log10 mfi to case/control
 anals <- unique(cyto_logit$analyte)
 gg_igg1 <- ggplot(cyto_logit[isotype=="IgG" & analyte %in% anals[1:21]],
                  aes(x=case, y=log10_mfi)) + 
   geom_boxplot() + xlab("case") + facet_wrap(~analyte) + ggtitle("IgG 1")
 
 gg_igg2 <- ggplot(cyto_logit[isotype=="IgG" & analyte %in% anals[22:42]],
                  aes(x=case, y=log10_mfi)) + 
   geom_boxplot() + xlab("case") + facet_wrap(~analyte) + ggtitle("IgG 2")
 
  gg_igm1 <- ggplot(cyto_logit[isotype=="IgM" & analyte %in% anals[1:21]],
                  aes(x=case, y=log10_mfi)) + 
   geom_boxplot() + xlab("case") + facet_wrap(~analyte) + ggtitle("IgM 1")
 
 gg_igm2 <- ggplot(cyto_logit[isotype=="IgM" & analyte %in% anals[22:42]],
                  aes(x=case, y=log10_mfi)) + 
   geom_boxplot() + xlab("case") + facet_wrap(~analyte) + ggtitle("IgM 2")
 
 ## output p-values and plots
 pvals %>% 
   mutate(V1=signif(V1, 4)) %>%
   dplyr::rename("p-value"=V1) %>%
   spread(key="isotype", value="p-value") %>%
   mutate(IgG_fdr=signif(p.adjust(IgG, method="fdr"), 4)) %>%
   mutate(IgM_fdr=signif(p.adjust(IgM, method="fdr"), 4)) %>%
   mutate(IgG=cell_spec(IgG, color = ifelse(IgG <= 0.2, "red", "black"))) %>%
   mutate(IgM=cell_spec(IgM, color = ifelse(IgM <= 0.2, "red", "black"))) %>%
   mutate(IgG_fdr=cell_spec(IgG_fdr, color = ifelse(IgG_fdr <= 0.2, 
                                                    "red", "black"))) %>%
   mutate(IgM_fdr=cell_spec(IgM_fdr, color = ifelse(IgM_fdr <= 0.2, 
                                                    "red", "black"))) %>%
 headerKable(title="clogit p-values")
```
 
##Log10 ratio plots of case and control
 
```{r plots} 
 print(gg_igg1)
 print(gg_igg2)
 print(gg_igm1)
 print(gg_igm2)
```
  
#GSEA analysis {.tabset .tabset-fade .tabset-pills}

```{r voom_setup}

fdr_cut <- 0.2

## reformat antibody data for limma
cyto %>%
 mutate(analyte= str_replace_all(analyte, "_", ".")) %>%
 mutate(antibody=paste(isotype, analyte, sep="_")) %>%
 mutate(antibody= str_replace_all(antibody, "[()]", "")) ->
    cyto_tmp

## get vector of unique combinations of isotype and analyte
antibodies <- sort(unique(cyto_tmp$antibody))

cyto_tmp %>%
 dcast(pid+col_id+visit+vaccine+case+malaria_status+stimulation+plate~antibody,
       value.var="log10_mfi") ->
  cyto_df

anti_voom <- mal067_voom[, cyto_df$col_id]
anti_voom$targets <- cyto_df
if(sum(colnames(anti_voom) != anti_voom$targets$col_id) > 0) {
  stop("GSEA join is screwed up")
}

applyGSEA <- function(antibody, dat, btm=btm_ind) {
  
  ## remove any potential NAs
  na_ind <- is.na(dat$targets[,antibody])
  voomy <- dat[,!na_ind]
  
  form <- paste0("~plate + ", antibody)
  des <- model.matrix(formula(form), voomy$targets)
  colnames(des)[ncol(des)] <- "antibody"
 
  ## run camera
  cam_vac <- as_tibble(camera(voomy, btm, design=des,
                              contrast="antibody"),
                       rownames="geneset")
  cam_vac %<>%
    mutate(PValue=signif(PValue, 3)) %<>%
    mutate(FDR=signif(FDR, 3))
  
  fit1 <- lmFit(voomy, des)
  fit2 <- eBayes(fit1, trend=FALSE)
  tmp_pfs <- topTable(fit2, number=Inf, coef="antibody", sort="P")
 
  return(cam_vac)
}

```
 
 
```{r samples}
anti_voom$targets %>%
  countSampleSizes("visit", c("vaccine", "malaria_status", "stimulation")) %>%
  headerKable(labels=c("vaccine", "malaria", "stim"), levels=c(1,2,1))
 
```
 
There are only 4 samples for M0 visit so that time period will not be analysed.  
  
##M3  
  
```{r voom}

## select only M3 samples
anti_voom$targets %>%
  filter(visit=="M3") %$%
  col_id ->
  m3_ind

m3_voom <- anti_voom[,m3_ind]

all_cam <- lapply(antibodies, applyGSEA, dat=m3_voom)
names(all_cam) <- antibodies

## get number of signficant gene sets
num_sigs <- sapply(all_cam, function(x) return(sum(x$FDR<=fdr_cut)),
                   simplify=TRUE)
out_sigs <- data.frame('significant sets'=num_sigs)

splits <- str_split(rownames(out_sigs), "_", simplify=TRUE)
colnames(splits) <- c("isotype", "analyte")
out_sigs <- cbind(out_sigs, splits)
out_all <- dcast(out_sigs, analyte~isotype, value.var="significant.sets")

## get significant gene sets
sigs <- lapply(all_cam, function(x) return(x[x$FDR <= fdr_cut,]))

```
  
###Summary of signficant sets  
  
```{r out_csp}
headerKable(out_all, title="Number of significant sets")
```
  
###Tables of significant sets  {.tabset .tabset-fade .tabset-pills}
  
```{r tables}
for(name in rownames(out_sigs)) {
  tmp <- sigs[[name]]
  if(nrow(tmp) > 0) {
     knit_print.html(cat(paste("####", name, "{-} \n  \n")))
     knit_print.html(cat(headerKable(tmp), "\n"))
  }
}

```

```{r deg}

applyDEG <- function(antibody, dat) {
  
  colInd <- c("logFC", "P.Value", "adj.P.Val")
  
  ## remove any potential NAs
  na_ind <- is.na(dat$targets[,antibody])
  voomy <- dat[,!na_ind]
  
  form <- paste0("~plate + ", antibody)
  des <- model.matrix(formula(form), voomy$targets)
  colnames(des)[ncol(des)] <- "antibody"
 
  fit1 <- lmFit(voomy, des)
  fit2 <- eBayes(fit1, trend=FALSE)
  tmp_pfs <- topTable(fit2, number=Inf, coef="antibody", sort="P")

 
}

all_cam <- lapply(antibodies, applyDEG, dat=m3_voom)
names(all_cam) <- antibodies

applyDEG(antibodies[1], dat=m3_voom)

```


