---
title: "Integrate IAVI and RNASeq data"
subtitle: "M3"
author: "Carl Murie and Raphael Gottardo"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
   BiocStyle::html_document:
    theme: lumen
    toc: true
    toc_float: true
    number_sections: true
---

```{r overall-knitr-options, echo=FALSE, results='hide'}
library(knitr)
opts_chunk$set(cache=FALSE, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results='asis')
```

```{r packages}
library(RNASeqUtilities)
library(Biobase)
library(limma)
library(magrittr)
library(here)
library(data.table)
library(survival)   ## clogit
library(GSEABase)
library(tidyverse)
library(DT)

source(here("code", "mal_utilities.R"))
```

```{r load-data}
library(mal067data)
data(mal067_eset)
meta_dt <- as_tibble(pData(mal067_eset))
```


```{r gsea}
min_gene_set <- 5

## set up GSEA analysis
btm_gtm <- getGmt(here("data/BTM_for_GSEA_20131008.gmt"))
btm_ind <- ids2indices(geneIds(btm_gtm), rownames(mal067_eset))
btm_ind <- btm_ind[sapply(btm_ind, length) > min_gene_set]
```


```{r getmorevars}

## l_df_flow and l_df_rna
##
## 1. cytokines: stimulation (HBS, CSP),antigens (28), and log10 stim/DMSO ratios
## 2. IAVI: IgG (CSP - NANP), IgG (CSP - C-term)
## 3. CEVAC: IgG CSP-NANP titers, IgG HBsAg titers (both in log10 of the titers)
## 4. antibody Luminex: isotope (IgG, IgM), analyte (42 antigens), log10 of the
##                      median fluorescence intensity
load("data/m067_seattle_data.RData")

```

# IAVI and RNASeq analysis  
  
Database of IAVI contains 2 new variables: eu_elisa_total.nanp (IgG CSP-NANP) and eu_elisa_total.cterm (IgG CSP-C-term). Both concentrations are given in European Units (EU) per mililiter. 
  
```{r iavi} 

## format rnaseq meta data
meta_dt %>%
  filter(stimulation=="csp") %>%
  filter(vaccine=="rtss") %>%
  mutate(visit=as.character(visit)) %>%
  dplyr::select(pid, visit, col_id, case, malaria_status, match,
                vaccine, stimulation, plate) ->
  meta_cyto

## format IAVI data
l_df_rna[[2]] ->
  df_rna

##countSampleSizes(df_rna, "visit", c("vaccine", "sex")) %>%
##  headerKable()

## join iavi data with rnaseq meta data
dplyr::inner_join(df_rna, meta_cyto, by=c("pid", "visit")) %>%
  dplyr::rename(nanp = eu_elisa_total.nanp) %>%
  dplyr::rename(cterm = eu_elisa_total.cterm) ->
  cyto

## check that total sample size is correct
total <- length(intersect(paste(meta_cyto$visit,meta_cyto$pid),
                          paste(df_rna$visit, df_rna$pid))) 
if(total != nrow(cyto)) {stop("Sample size is not correct")}


```

## Sample size

### IAVI database  
  
```{r samplesize1}
countSampleSizes(df_rna, "visit", c("vaccine", "site")) %>%
  headerKable()
```

### after joining the rnaSeq meta-data and IAVI database  
  
Only rtss samples were used.  

```{r samplesize2}
countSampleSizes(cyto, "visit", c("vaccine.x", "case", "stimulation")) %>%
  headerKable()
```

One sample(pid=3010 and visit=M0) in the AIVI database was identified as given the comparator while the RNASeq meta-data had it assigned as being given the rts,s vaccine. This sample was removed.  
  
```{r clean}
cyto %<>%
  filter(pid !=3010)
```

## Disease prediction  

A clogit test of case-control vs isotype differences with gender and match as a stratified variable was applied to the IAVI frequencies.  

```{r cyto_logit}

## gather just case/control samples which the match stratified variable
cyto %>%
  filter(case != "neither") %>%
  mutate(case = fct_drop(case)) %>%                    ## remove "neither" level
  ## clogit needs case as a binary value (0,1)
  mutate(case_logit=as.numeric(plyr::mapvalues(as.character(case), 
                                               from=c("control", "case"),
                                               to=c(0,1)))) %>%
  as.data.table() ->
  cyto_logit
```

###Sample size  
  
```{r sample3} 
countSampleSizes(cyto_logit, "visit", c("case", "vaccine.x")) %>%
  headerKable()
```

```{r output}

 ##x <- clogit(case_logit~sex+cterm+strata(match), data=cyto_logit[visit=="M0"])

 pval_cterm <- cyto_logit[,summary(clogit(case_logit~sex+cterm+strata(match)))$coefficient["cterm", 5], by=.(visit)]
colnames(pval_cterm)[2] <- "cterm"

pval_nanp <- cyto_logit[,summary(clogit(case_logit~sex+nanp+strata(match)))$coefficient["nanp", 5], by=.(visit)]
colnames(pval_nanp)[2] <- "nanp"
 
  ## plot boxplots of log10 ratio to case/control
 gg_cterm <- ggplot(cyto_logit, aes(x=case, y=cterm)) + geom_point() + 
   geom_jitter() + xlab("case") + facet_wrap(~visit) + ggtitle("cterm")
 
 gg_nanp <- ggplot(cyto_logit, aes(x=case, y=nanp)) + geom_point() + 
   geom_jitter() + xlab("case") + facet_wrap(~visit) + ggtitle("nanp")
```
  
The clogit function did not converge for the nanp data.  
  
```{r table}
 ## output p-values and plots
 cbind(pval_nanp, pval_cterm[,2]) %>%
   mutate(nanp=signif(nanp, 4)) %>%
   mutate(cterm=signif(cterm, 4)) %>%
   mutate(nanp_fdr=signif(p.adjust(nanp, method="fdr"), 4)) %>%
   mutate(cterm_fdr=signif(p.adjust(cterm, method="fdr"), 4)) %>%
   mutate(nanp=cell_spec(nanp, color = ifelse(nanp <= 0.2, "red", "black"))) %>%
   mutate(cterm=cell_spec(cterm, color = ifelse(cterm <= 0.2, "red", "black"))) %>%
   mutate(nanp_fdr=cell_spec(nanp_fdr, color = ifelse(nanp_fdr <= 0.2, 
                                                      "red", "black"))) %>%
   mutate(cterm_fdr=cell_spec(cterm_fdr, color = ifelse(cterm_fdr <= 0.2, 
                                                        "red", "black"))) %>%
 headerKable(title="clogit p-values")
```
 
##nanp and cterm plots of case and control
 
```{r plots} 
 print(gg_nanp)
 print(gg_cterm)
```
  
##GSEA analysis {.tabset .tabset-fade .tabset-pills}

```{r voom}
iavi_voom <- mal067_voom[, cyto$col_id]
iavi_voom$targets <- cyto
if(sum(colnames(iavi_voom) != iavi_voom$targets$col_id) > 0) {
  stop("GSEA join is screwed up")
}

## model definition
form_nanp <- "~plate + nanp"
coef_nanp <- "nanp"
form_cterm <- "~plate + cterm"
coef_cterm <- "cterm"
fdr_cut <- 0.2
```

####Sample Size  
  
```{r samps}

cyto %>%
  countSampleSizes("stimulation", c("vaccine.x", "visit")) %>%
  headerKable()

```

linear model: `r form_nanp`  
coefficient: `r coef_nanp`  
  
linear model: `r form_cterm`  
coefficient: `r coef_cterm`  
  
FDR cutoff <= `r fdr_cut`   

###M0  {.tabset .tabset-fade .tabset-pills}
  
```{r m0} 

## filter for m0 time point
cyto %>%
  filter(visit == "M0") %$%
  col_id ->
  cyto_ind 

voom_m0 <- iavi_voom[,cyto_ind]
```
  
####NANP  
  
```{r nanp}

## remove NAs
## remove duplicate
voom_m0$targets %>%
  filter(!is.na(nanp))%$%
  col_id ->
  clean_ind

voom_nanp <- voom_m0[,clean_ind]

## run camera
des <- model.matrix(formula(form_nanp), voom_nanp$targets)
cam <- as_tibble(camera(voom_nanp, btm_ind, design=des,
                              contrast=coef_nanp),
                       rownames="geneset")
 cam %>%
    mutate(PValue=signif(PValue, 3)) %>%
    mutate(FDR=signif(FDR, 3)) %>%
   filter(FDR <= fdr_cut) %>%
   datatable()
 
 sum_m0 <- sum(cam$FDR <= fdr_cut)

```
  
####CTERM  
  
```{r cterm}

## remove NAs
## remove duplicate
voom_m0$targets %>%
  filter(!is.na(cterm))%$%
  col_id ->
  clean_ind

voom_cterm <- voom_m0[,clean_ind]

## run camera
des <- model.matrix(formula(form_cterm), voom_cterm$targets)
cam <- as_tibble(camera(voom_cterm, btm_ind, design=des,
                              contrast=coef_cterm),
                       rownames="geneset")
 cam %>%
    mutate(PValue=signif(PValue, 3)) %>%
    mutate(FDR=signif(FDR, 3)) %>%
   filter(FDR <= fdr_cut) %>%
   datatable()
 
 sum_m0 <- c(sum_m0, sum(cam$FDR <= fdr_cut))

```
  
###M3  {.tabset .tabset-fade .tabset-pills}
  
```{r m3} 

## filter for m0 time point
cyto %>%
  filter(visit == "M3") %$%
  col_id ->
  cyto_ind 

voom_m3 <- iavi_voom[,cyto_ind]
```
  
####NANP  
  
```{r nanp_m3}

## remove NAs
## remove duplicate
voom_m3$targets %>%
  filter(!is.na(nanp))%$%
  col_id ->
  clean_ind

voom_nanp <- voom_m3[,clean_ind]

## run camera
des <- model.matrix(formula(form_nanp), voom_nanp$targets)
cam <- as_tibble(camera(voom_nanp, btm_ind, design=des,
                              contrast=coef_nanp),
                       rownames="geneset")
 cam %>%
    mutate(PValue=signif(PValue, 3)) %>%
    mutate(FDR=signif(FDR, 3)) %>%
   filter(FDR <= fdr_cut) %>%
   datatable()

sum_m3 <- sum(cam$FDR <= fdr_cut) 
```
  
####CTERM  
  
```{r cterm_m3}

## remove NAs
## remove duplicate
voom_m3$targets %>%
  filter(!is.na(cterm))%$%
  col_id ->
  clean_ind

voom_cterm <- voom_m3[,clean_ind]

## run camera
des <- model.matrix(formula(form_cterm), voom_cterm$targets)
cam <- as_tibble(camera(voom_cterm, btm_ind, design=des,
                              contrast=coef_cterm),
                       rownames="geneset")
 cam %>%
    mutate(PValue=signif(PValue, 3)) %>%
    mutate(FDR=signif(FDR, 3)) %>%
   filter(FDR <= fdr_cut) %>%
   datatable()

sum_m3 <- c(sum_m3, sum(cam$FDR <= fdr_cut))
```
  
##Summary  
  
```{r summary}
sum <- rbind(sum_m0, sum_m3)
dimnames(sum) <- list(c("M0", "M3"), c("NANP", "CTERM"))

headerKable(sum)
```

