---
title: "RTSS GSEA ~ monocyte report"
author: "Carl Murie"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
 html_document:
    theme: lumen
    toc: true
    toc_float: true
    number_sections: true
    df_print: paged
    code_folding: hide
---

```{r overall-knitr-options, cache=FALSE, echo=FALSE, message=FALSE, warning=FALSE, results="hide"}
library(knitr)

opts_chunk$set(cache=FALSE, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE,
               results='asis')
  
  library(dplyr)
  library(RNASeqUtilities)
  library(data.table)
  library(Biobase)
  library(edgeR)
  library(GSEABase)
  library(DT)
  library(limma)
  library(ggplot2)
  library(gridExtra)
  library(kableExtra)
  library(tidyverse)
  library(survival)    ## clogit
  library(lme4)        ## glmer
```


```{r load-data}
library(mal067data)
data(mal067_meta_flow)

mal067_meta_master <- rbind(mal067_meta_flow[,c("sid", "visit", "vaccine", "case", "match", "sex")], mal067_meta_rnaseq[,c("sid", "visit", "vaccine", "case", "match", "sex")])

mal067_meta_master %>%
  filter(!duplicated(sid)) %>%
  dplyr::select(c(sid, visit, vaccine, case, match, sex)) ->
  mal_pheno

## deRosa phenotype frequencies
flow <- read.table(here::here("data/phenoLineagePctLive.txt"),
          sep="\t", stringsAsFactors=FALSE, header=TRUE,
          row.names=NULL)[,c("PTID","VISITNO","Lineage","Label","Pct.of.live")] 

flow %>%
setnames(c("Lineage", "Pct.of.live"), c("Population", "Frequency")) %>%
mutate(Frequency=sqrt(Frequency/100))  ->         ## convert to proportion 
         flowData1

## PTID "FH12 (SAC207295)" was discarded as there no samples in meta_master. I 
## assume these are some kind of control.

## Also PTID (sid) 5094082 was assigned visit M0 in the flow phenotype data and 
## assigned visit M3 in the meta_master data. I have contacted Barcelona for 
## clarification but for now I am just removing it.
dplyr::inner_join(flowData1, mal_pheno, by=c("PTID"="sid")) %>%
  ##filter(PTID != "5094082") %>%
  mutate(match=as.numeric(match)) ->
  flowData
```

# DeRosa phenotype cell subset frequencies  
  
##Sample sizes
  
```{r case_sample}
countSampleSizes(flowData[!duplicated(flowData$PTID),], "visit", 
                 c("case", "vaccine")) %>%
  headerKable(labels=c("disease", "vaccine"), levels=c(3,2))
```
  
# Case vs control for all phenotype populations  
  
A clogit test of case-control vs frequency (on sqrt scale) with gender and match as a stratified variable was applied to the cell frequency. 

##Sample sizes  
  
Per cell population
  
```{r case}

flowData %>%
  filter(case != "neither") %>%
  mutate(case=fct_drop(case)) %>%
  filter(vaccine == "rtss") %>%
  mutate(case_logit=as.numeric(plyr::mapvalues(case, 
                                               from=c("control", "case"),
                                               to=c(0,1)))) %>%   
  data.table() -> 
  case_data

countSampleSizes(case_data[Population=="CD19+"], "visit", c("case", "vaccine", "sex")) %>%
  headerKable(labels=c("disease", "vaccine", "gender"), levels=c(2,2,2))
```
 
## clogit pValues  
  
```{r disease}

logit_pvals <- case_data[, signif( summary(clogit(case_logit~Frequency+sex+strata(match)))$coefficients["Frequency", 5], 3), by=.(visit, Population)]

logit_pvals %>%
  spread(key=visit, value=V1) %>%
  mutate(M0_fdr=signif(p.adjust(M0, method="fdr"), 3)) %>%
  mutate(M3_fdr=signif(p.adjust(M3, method="fdr"), 3)) %>%
  mutate(M0=cell_spec(M0, color = ifelse(M0 <= 0.2, "red", "black"))) %>%
  mutate(M3=cell_spec(M3, color = ifelse(M3 <= 0.2, "red", "black"))) %>%
  mutate(M0_fdr=cell_spec(M0_fdr, color = ifelse(M0_fdr<=0.2, "red", "black"))) %>%
  mutate(M3_fdr=cell_spec(M3_fdr, color = ifelse(M3_fdr<=0.2, "red", "black"))) ->
  pvals

 headerKable(pvals)
```
  
## Plots of square root of cell subset frequency  
  
```{r ggplot}

 ## plot boxplots of log10 ratio to case/control
 ggplot(case_data[visit=="M0"], aes(x=case, y=Frequency)) + 
   geom_boxplot() + xlab("case") + facet_wrap(~Population) + ggtitle("M0")

 ggplot(case_data[visit=="M3"], aes(x=case, y=Frequency)) + 
   geom_boxplot() + xlab("case") + facet_wrap(~Population) + ggtitle("M3")

```

# Vaccine effect for all phenotype populations  
  
A clogit test of vaccine vs frequency (on sqrt scale) with gender and match as a stratified variable was applied to the cell frequency.  
  
This includes all case/control subjects and both rtss and comparator but only visit M3.

##Sample sizes  
  
Per cell population
  
```{r vacc_sample}

flowData %>%
  filter(visit=="M3") %>%
  mutate(vaccine=as.character(vaccine)) %>%
  mutate(vacc_logit=as.numeric(plyr::mapvalues(vaccine, 
                                               from=c("comparator", "rtss"),
                                               to=c(0,1)))) %>%   
  data.table() -> 
  vacc_data

countSampleSizes(vacc_data[Population=="CD19+"], "vaccine", c("case", "sex")) %>%
  headerKable(labels=c("disease", "gender"), levels=c(3,2))
```
 
## clogit pValues  
  
```{r vaccine}

##logit_pvals <- vacc_data[, signif( summary(clogit(vacc_logit~Frequency+sex+strata(match)))$coefficients["Frequency", 5], 3), by=.(Population)]

logit_pvals <- vacc_data[, signif(summary(glmer(vacc_logit~Frequency+sex+(1|match), family="binomial"))$coefficients["Frequency", 4], 3), by=.(Population)]

colnames(logit_pvals)[2] <- "pValue"

logit_pvals %>%
  mutate(fdr=signif(p.adjust(pValue, method="fdr"), 3)) %>%
  mutate(pValue=cell_spec(pValue, color = ifelse(pValue<=0.2, "red", "black"))) %>%
  mutate(fdr=cell_spec(fdr, color = ifelse(fdr<=0.2, "red", "black"))) ->
  pvals

 headerKable(pvals)
```
  
## Plots of square root of cell subset frequency  
  
```{r ggplot2}

 ## plot boxplots of log10 ratio to case/control
 ggplot(vacc_data, aes(x=vaccine, y=Frequency)) + 
   geom_boxplot() + xlab("vaccine") + facet_wrap(~Population) + ggtitle("M3")

```
