---
title: "Integrate CEVAC and RNASeq data"
author: "Carl Murie and Raphael Gottardo"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: 
   BiocStyle::html_document:
    theme: lumen
    toc: true
    toc_float: true
    number_sections: true
---

```{r overall-knitr-options, echo=FALSE, results='hide'}
library(knitr)
opts_chunk$set(cache=FALSE, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results='asis')
```

```{r packages}
library(RNASeqUtilities)
library(Biobase)
library(limma)
library(magrittr)
library(here)
library(data.table)
library(survival)   ## clogit
library(GSEABase)
library(tidyverse)
library(DT)

source(here("code", "mal_utilities.R"))
```

```{r load-data}
library(mal067data)
data(mal067_eset)
meta_dt <- as_tibble(pData(mal067_eset))
```


```{r gsea}
min_gene_set <- 5

## set up GSEA analysis
btm_gtm <- getGmt(here("data/BTM_for_GSEA_20131008.gmt"))
btm_ind <- ids2indices(geneIds(btm_gtm), rownames(mal067_eset))
btm_ind <- btm_ind[sapply(btm_ind, length) > min_gene_set]
```


```{r getmorevars}

## l_df_flow and l_df_rna
##
## 1. cytokines: stimulation (HBS, CSP),antigens (28), and log10 stim/DMSO ratios
## 2. IAVI: IgG (CSP - NANP), IgG (CSP - C-term)
## 3. CEVAC: IgG CSP-NANP titers, IgG HBsAg titers (both in log10 of the titers)
## 4. antibody Luminex: isotope (IgG, IgM), analyte (42 antigens), log10 of the
##                      median fluorecence intensity
load("data/m067_seattle_data.RData")

```

# CEVAC and RNASeq analysis  
  
Database of CEVAC contains 2 new variables: log_val.Anti-CS (IgG CSP-NANP titers) and  log_val.HBV.S AB (IgG HBsAg titers), as log10 of the titers.

csp: IgG CSP-NANP titers  
hbs: IgG HBsAg titers  
   
```{r iavi} 

## format rnaseq meta data
meta_dt %>%
  ##filter(stimulation=="csp" | stimulation=="hbs") %>%
  filter(stimulation=="dmso") %>%
  mutate(stimulation = as.character(fct_drop(stimulation))) %>%
  filter(vaccine=="rtss") %>%
  mutate(visit=as.character(visit)) %>%
  dplyr::select(pid, visit, col_id, case, malaria_status,
                match, vaccine, plate, total_reads, 
                age_weeks, sex) ->
  meta_cyto

## format CEVAC data
l_df_rna[[3]] %>%
  dplyr::rename(CSP_NANP="log_val.Anti-CS") %>%
  dplyr::rename(HBsAg= "log_val.HBV.S AB") %>%
  melt(measure=c("CSP_NANP", "HBsAg")) %>%
  dplyr::rename(stimulation=variable) %>%
  dplyr::rename(log10_score=value) %>%
  filter(!is.na(log10_score)) %>%
  mutate(stimulation=as.character(stimulation))->
  df_rna
```

##Sample size  
  
```{r}
countSampleSizes(df_rna, "visit", "stimulation") %>%
  headerKable()
```

##Plots of new variables

```{r plot}
ggplot(df_rna, aes(x=visit, y=log10_score)) +
  geom_boxplot() + geom_point() + geom_jitter() +
  facet_wrap(~stimulation, scales="free") + xlab("order")
```

```{r}
## join cevac data with rnaseq meta data
dplyr::inner_join(df_rna, meta_cyto, by=c("pid", "visit")) ->
  cyto
```

## Sample size  
  
There are only 3 samples at time point M0 after joining the RNASeq and titers data so only the M3 time point will be analysed.
  
```{r samplesize1}
 countSampleSizes(cyto, "visit", 
                  c("stimulation", "vaccine", "case")) %>%
  headerKable()
```

## Disease prediction  

A clogit test of case-control vs isotype titers with gender and match as a stratified variable was applied to the CEVAC measurements.  

```{r cyto_logit}

## gather just case/control samples which the match stratified variable
cyto %>%
  filter(case != "neither") %>%
  mutate(case = fct_drop(case)) %>%                    ## remove "neither" level
  ## clogit needs case as a binary value (0,1)
  mutate(case_logit=as.numeric(plyr::mapvalues(as.character(case), from=c("control", "case"), to=c(0,1)))) %>%
  filter(visit=="M3") %>%
  mutate(visit = fct_drop(visit)) %>%
  as.data.table() ->
  cyto_logit
```

###Sample size  
  
```{r sample3} 
countSampleSizes(cyto_logit, "stimulation", c("case", "vaccine")) %>%
  headerKable()
```

```{r output}
 ##x <- clogit(case_logit~sex+cterm+strata(match), data=cyto_logit[visit=="M0"])

 pvals <- cyto_logit[,summary(clogit(case_logit~sex+log10_score+strata(match)))$coefficient["log10_score", 5], by=.(stimulation)]
colnames(pvals)[2] <- "pValue"
 
  ## plot boxplots of log10 ratio to case/control
 gg_plott <- ggplot(cyto_logit, aes(x=case, y=log10_score)) + geom_boxplot() +
   xlab("case") + facet_wrap(~stimulation)
 
```
  
  
```{r table}
 ## output p-values and plots
 pvals %>%
   mutate(pValue=signif(pValue, 4)) %>%
   mutate(FDR=signif(p.adjust(pValue, method="fdr"), 4)) %>%
   mutate(FDR=cell_spec(FDR, color = ifelse(FDR<=0.2, "red", "black"))) %>%
   mutate(pValue=cell_spec(pValue, color = ifelse(pValue<=0.2, "red", "black"))) %>%
 headerKable(title="clogit p-values")
```
 
##plots of case and control
 
```{r plots} 
 print(gg_plott)
```
  
#GSEA analysis

The titers were drawn from only DMSO samples.  
  
The M0 visit will not be analysed as there are only 3 samples.   

```{r gsea_setup}
cyto %>%
  filter(visit == "M3") %>%
  dplyr::rename(titers="stimulation") ->
  cyto_m3 

voom_gsea <- mal067_voom[,cyto_m3$col_id]
targ <- voom_gsea$targets
voom_gsea$targets <- cyto_m3
if(sum(colnames(voom_gsea) != voom_gsea$targets$col_id) > 0) {
  stop("GSEA titer join is screwed up")
}

## model definition
form <- "~plate + total_reads + age_weeks + titers"
coef <- "log10_score"
fdr_cut <- 0.2
```
  
linear model: `r form`  
coefficient: `r coef`  
FDR cutoff <= `r fdr_cut`  
  
##Sample sizes  
  
```{r samps}
countSampleSizes(voom_gsea$targets, "titers", 
            c("visit", "vaccine", "malaria_status")) %>%
  headerKable(c("visit", "vaccine", "disease"),
              levels=c(1,2,2))
```

##M3 {.tabset .tabset-fade .tabset-pills}

###IgG CSP-NANP titers  
  
```{r csp}
csp_ind <- voom_gsea$targets$titers=="CSP_NANP"
voom_csp <- voom_gsea[,csp_ind]

 form <- "~plate + total_reads + age_weeks + log10_score"
 des <- model.matrix(formula(form), voom_csp$targets)
 
## run camera
 cam_vac <- as_tibble(camera(voom_csp, btm_ind, design=des,
                              contrast=coef),
                       rownames="geneset")
 cam_vac %>%
    mutate(PValue=signif(PValue, 3)) %>%
    mutate(FDR=signif(FDR, 3)) %>%
   filter(FDR <= fdr_cut) %>%
   datatable()

 summ <- sum(cam_vac$FDR <= fdr_cut)
 
```
  
###IgG HBsAg titers  
  
```{r hbs}
csp_ind <- voom_gsea$targets$titers=="HBsAg"
voom_csp <- voom_gsea[,csp_ind]

 form <- "~plate + total_reads + age_weeks + log10_score"
 des <- model.matrix(formula(form), voom_csp$targets)
 
## run camera
 cam_vac <- as_tibble(camera(voom_csp, btm_ind, design=des,
                              contrast=coef),
                       rownames="geneset")
 cam_vac %>%
    mutate(PValue=signif(PValue, 3)) %>%
    mutate(FDR=signif(FDR, 3)) %>%
   filter(FDR <= fdr_cut) %>%
   datatable()

 summ <- c(summ, sum(cam_vac$FDR <= fdr_cut))
```  
  
##Summary
  
```{r summ}
outie <- data.frame(number=summ)
rownames(outie) <- c("CSP-NANP", "HBsAg")
headerKable(outie, title="number of significant gsea sets")
```
