---
title: 'Manuscript: M3 DMSO'
author: "Carl Murie"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r overall-knitr-options, echo=FALSE, results='hide'}
library(knitr)
opts_chunk$set(cache=FALSE, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results='asis')
```

```{r packages}
library(readr)
library(RNASeqUtilities)
library(Biobase)
library(magrittr)
library(here)
library(heatmap3)
library(DT)
library(GSEABase)
library(limma)
library(data.table)
library(RColorBrewer)
library(heatmap3)
library(tidyverse)

source(here("code", "mal_utilities.R"))
```

```{r load}
## load mal067 data
library(mal067data)
meta_dt <- as_tibble(pData(mal067_eset))

DO_PDF <- FALSE

stim="dmso"
time="M3"
```

```{r genesets}
## set up GSEA analysis
min_gene_set <- 5
btm_gtm <- getGmt(here("data/BTM_for_GSEA_20131008.gmt"))
btm_ind <- ids2indices(geneIds(btm_gtm), rownames(mal067_eset))
btm_ind <- btm_ind[sapply(btm_ind, length) > min_gene_set]

## read in BTM module annotations
btm_anno <- read_csv(here("data/btm_annotation_table_LNC.csv"))
btm_anno$full_title <- paste0(btm_anno$`Module title`, " (", btm_anno$ID, ")")

```

```{r set-parameters}

## parameters set by parent: stim, time

fdr_cut <- 0.2
p_cut <- 0.01
form_main_dis_case <- "~plate + total_reads + age + case"
coeff_case <- "casecase"
##form_main_dis_malaria <- "~plate + total_reads + age + malaria_status"
##coeff_malaria <- "malaria_status1"
##coeff <- "casecase"
```

 
# GSEA `r toupper(stim)`: disease main effects {.tabset .tabset-fade .tabset-pills}

**disease linear model:** `r form_main_dis_case`  
**disease filtering   :** `r time`

```{r disease_data}

## disease: select subset of expressionSet
meta_dt %>% 
  filter(stimulation == stim) %>% 
  filter(case == "case" | case == "control") %>%
  filter(visit == time) %$%
  col_id ->
  sample_vac

## generate subset data and drop extra stimulation/case levels
mal_dis <- mal067_eset[, sample_vac] 
mal_dis$stimulation <- fct_drop(mal_dis$stimulation)
mal_dis$case <- fct_drop(mal_dis$case)
```


```{r sample_size}

## get sample sizes
countSampleSizes(pData(mal_dis), "stimulation", 
            c("vaccine", "malaria_status", "age", "visit", "case")) %>%
headerKable(labels=c("vaccine", "malaria", "age", "visit", "case"),
            levels=c(2,2,2,2,2), title="disease sample size")

```
  
##disease M3 (malaria - no malaria): RTS,S only

```{r dis_analysis_rtss}
################## vaccine analysis ########################

cam_dis_rtss <- runGSEA(mal_dis[,mal_dis$vaccine=="rtss"], 
                   form_main_dis_case,
                   coef=coeff_case)
numSig <- sum(cam_dis_rtss$FDR <= fdr_cut)

suffix <- paste(stim, time, sep="_")
write_csv(cam_dis_rtss, paste0(here("output/"), "disease_rtss_", suffix, ".csv"))

```
  
##disease M3 (malaria - no malaria): comparator only  
  
```{r dis_analysis_comp}
################## vaccine analysis ########################

cam_dis_comp <- runGSEA(mal_dis[,mal_dis$vaccine=="comparator"], 
                   form_main_dis_case,
                   coef=coeff_case)
numSig <- sum(cam_dis_comp$FDR <= fdr_cut)

suffix <- paste(stim, time, sep="_")
write_csv(cam_dis_comp, paste0(here("output/"), "disease_comp_", suffix, ".csv"))

```
  
  
```{r deg_mal}
 colInd <- c("logFC", "P.Value", "adj.P.Val")

 mal_rtss <- mal_dis[, mal_dis$vaccine=="rtss"]
 design_dis <- model.matrix(formula(form_main_dis_case), mal_rtss)
 mal_voom <- voom(mal_rtss, design=design_dis)

 fit1 <- lmFit(mal_voom, design_dis)
 fit2 <- eBayes(fit1, trend=FALSE)
 tmp_deg <- topTable(fit2, number=Inf, coef="casecase", sort="P")
 vc_deg <- tmp_deg
 vc_deg$genenames <- rownames(vc_deg)


```
  
  
#Plots: M3 DMSO  
  
```{r heatme_vac}

## reformat and process
cam_dis_rtss %>%
  dplyr::select(-c(NGenes)) %>%
  dplyr::left_join(cam_dis_comp, by="geneset") %>%
  dplyr::select(-c(NGenes)) %>%
  dplyr::left_join(btm_anno, by=c("geneset"="full_title")) %>% 
  filter(!is.na(annotation) & !str_detect(annotation,"TBA")) %>%
  dplyr::rename(PValue_rtss=PValue.x,
                Direction_rtss=Direction.x,
                FDR_rtss=FDR.x,
                PValue_comp=PValue.y,
                Direction_comp=Direction.y,
                FDR_comp=FDR.y) %>%
  mutate(Direction_rtss=as.numeric(plyr::mapvalues(Direction_rtss, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(Direction_comp=as.numeric(plyr::mapvalues(Direction_comp, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(log10_Pval_rtss=Direction_rtss*-log10(PValue_rtss)) %>%
  mutate(log10_Pval_comp=Direction_comp*-log10(PValue_comp)) ->
  outie_m3

## select only significant gene sets
allsigs <- unique(c(cam_dis_rtss$geneset[cam_dis_rtss$FDR <= fdr_cut],
                    cam_dis_comp$geneset[cam_dis_comp$FDR <= fdr_cut]))

outie_m3%>%
   filter(geneset %in% allsigs) ->
outie


## set all non-significant genesets PValues to 1
outie$log10_Pval_rtss[outie$FDR_rtss >= fdr_cut] <- 0
outie$log10_Pval_comp[outie$FDR_comp >= fdr_cut] <- 0

outie %>%
  dplyr::arrange(annotation, log10_Pval_rtss, log10_Pval_comp) %>%
  dplyr::select(c(geneset, annotation, 
                  log10_Pval_rtss, 
                  log10_Pval_comp)) %>%
  column_to_rownames("geneset") %>%
  dplyr::rename(rtss=log10_Pval_rtss) %>%
  dplyr::rename(comp=log10_Pval_comp) %>%
  as.data.frame() ->
  heat_mat

## colors for annotation
annos <- unique(heat_mat$annotation)
anno_col <- rainbow(length(annos))
anno_cols <- anno_col[as.factor(heat_mat$annotation)]
heat_mat$annotation <- NULL

if(DO_PDF) {pdf(file=here("output/figures/manuscript/heat_disease_dmso_M3.pdf"))}

heatmap3(heat_mat, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, 
         cexRow=0.50, cexCol=1.25, showRowDendro=FALSE, margins=c(3,15),
         col=colorRampPalette(c("navy", "white", "firebrick3"))(1024),
         RowSideColors=anno_cols, RowSideLabs="")
 legend(x="bottomleft", inset=-0.11, legend=annos, fill=anno_col, bty="n", cex=0.7)
 
if(DO_PDF) {dev.off()}

```



##disease M3-MO  {.tabset .tabset-fade .tabset-pills}
  
Only Bagamoya samples with the dmso stimulation are used. A consequence of this is that only subjects from the old age group are included.  Additionally only paired samples by subject are included.
  
```{r m3-m0}

form_visit_main <- "~plate + total_reads + pid + visit"
coeff <- "visitM3"

mal067_voom$targets %>%
  filter(site=="BAGAMOYO") %>%
  filter(stimulation=="dmso") %>%
  filter(case == "case" | case == "control") %$%
  col_id ->
  bag_ind

v_bag <- mal067_voom[,bag_ind]
v_bag$targets$stimulation <- fct_drop(v_bag$targets$stimulation)
v_bag$targets$case <- fct_drop(v_bag$targets$case)
v_bag$targets$visit <- fct_relevel(v_bag$targets$visit, "M0")


## get paired intersection between time points
pids <- intersect(v_bag$targets$pid[v_bag$targets$visit=="M0"],
                  v_bag$targets$pid[v_bag$targets$visit=="M3"])
v_bag$targets %>%
  filter(pid %in% pids) %$%
  col_id ->
  pid_ids
  
v_bag <- v_bag[,pid_ids]

v_bag$targets %>%
  countSampleSizes("stimulation", c("vaccine", "malaria_status",  "age", "visit", "case")) %>%
  headerKable(labels=c("vaccine", "disease", "age", "visit", "case"),
              levels=c(2,2,2,2,2))
```

formula: `r form_visit_main`  
coefficient: `r coeff`  
  
###malaria and rtss
  
```{r visit_ma1l}

## select only rts,s subjects
indy <- v_bag$targets$malaria_status == 1 & v_bag$targets$vaccine=="rtss"

## interaction
des <- model.matrix(formula(form_visit_main), v_bag$targets[indy,])

## run camera
cam_visit_mal_rtss <- as_tibble(camera(v_bag$E[,indy], btm_ind, design=des,
                              contrast=coeff),
                       rownames="geneset")
v_mal_sigs <- sum(cam_visit_mal_rtss$FDR <= fdr_cut)

cam_visit_mal_rtss %>%
  filter(FDR <= fdr_cut) %>%
  datatable()

write_csv(cam_visit_mal_rtss, paste0(here("output/"),
                                 "M3-M0_disease_mal_rtss.csv"))

```

###malaria and comparator
  
```{r visit_comp}

## select only rts,s subjects
indy <- v_bag$targets$malaria_status == 1 & v_bag$targets$vaccine=="comparator"

## interaction
des <- model.matrix(formula(form_visit_main), v_bag$targets[indy,])

## run camera
cam_visit_mal_comp <- as_tibble(camera(v_bag$E[,indy], btm_ind, design=des,
                              contrast=coeff),
                       rownames="geneset")
v_mal_sigs <- sum(cam_visit_mal_comp$FDR <= fdr_cut)

cam_visit_mal_comp %>%
  filter(FDR <= fdr_cut) %>%
  datatable()

write_csv(cam_visit_mal_comp, paste0(here("output/"),
                                 "M3-M0_disease_mal_comp.csv"))

```



###no malaria and rtss
  
```{r visit_nomal}

## select only rts,s subjects
indy <- v_bag$targets$malaria_status == 0 & v_bag$targets$vaccine=="rtss"

## interaction
des <- model.matrix(formula(form_visit_main), v_bag$targets[indy,])

## run camera
cam_visit_nomal_rtss <- as_tibble(camera(v_bag$E[,indy], btm_ind, design=des,
                              contrast=coeff),
                       rownames="geneset")
v_mal_sigs <- sum(cam_visit_nomal_rtss$FDR <= fdr_cut)

cam_visit_nomal_rtss %>%
  filter(FDR <= fdr_cut) %>%
  datatable()

write_csv(cam_visit_nomal_rtss, paste0(here("output/"),
                                 "M3-M0_disease_nomal_rtss.csv"))

```

###no malaria and comparator
  
```{r visit_both}

## select only rts,s subjects
indy <- v_bag$targets$malaria_status == 0 & v_bag$targets$vaccine=="comparator"

## interaction
des <- model.matrix(formula(form_visit_main), v_bag$targets[indy,])

## run camera
cam_visit_nomal_comp <- as_tibble(camera(v_bag$E[,indy], btm_ind, design=des,
                              contrast=coeff),
                       rownames="geneset")
v_mal_sigs <- sum(cam_visit_nomal_comp$FDR <= fdr_cut)

cam_visit_nomal_comp %>%
  filter(FDR <= fdr_cut) %>%
  datatable()

write_csv(cam_visit_nomal_comp, paste0(here("output/"),
                                 "M3-M0_disease_nomal_comp.csv"))

```



#Plots: M3-M0    
  
```{r heatme_dis}

## filter and process malaria subjects
cam_visit_mal_rtss %>%
  dplyr::select(-c(NGenes)) %>%
  dplyr::left_join(cam_visit_mal_comp, by="geneset") %>%
  dplyr::select(-c(NGenes)) %>%
  dplyr::left_join(btm_anno, by=c("geneset"="full_title")) %>%
  dplyr::select(-c(ID, 'Module title')) %>%
  dplyr::rename(PValue_mal_r=PValue.x, 
                Direction_mal_r=Direction.x, 
                FDR_mal_r=FDR.x, 
                PValue_mal_c=PValue.y, 
                Direction_mal_c=Direction.y, 
                FDR_mal_c=FDR.y) %>% 
  mutate(Direction_mal_r=as.numeric(plyr::mapvalues(Direction_mal_r, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(Direction_mal_c=as.numeric(plyr::mapvalues(Direction_mal_c, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(log10_Pval_mal_r=Direction_mal_r*-log10(PValue_mal_r)) %>%
  mutate(log10_Pval_mal_c=Direction_mal_c*-log10(PValue_mal_c)) ->
  outie_m3_m0_mal

## filter and process no malaria subjects
cam_visit_nomal_rtss %>%
  dplyr::select(-c(NGenes)) %>%
  dplyr::left_join(cam_visit_nomal_comp, by="geneset") %>%
  dplyr::select(-c(NGenes)) %>%
  dplyr::rename(PValue_nomal_r=PValue.x, 
                Direction_nomal_r=Direction.x, 
                FDR_nomal_r=FDR.x, 
                PValue_nomal_c=PValue.y, 
                Direction_nomal_c=Direction.y, 
                FDR_nomal_c=FDR.y) %>% 
  mutate(Direction_nomal_r=as.numeric(plyr::mapvalues(Direction_nomal_r, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(Direction_nomal_c=as.numeric(plyr::mapvalues(Direction_nomal_c, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(log10_Pval_nomal_r=Direction_nomal_r*-log10(PValue_nomal_r)) %>%
  mutate(log10_Pval_nomal_c=Direction_nomal_c*-log10(PValue_nomal_c)) ->
  outie_m3_m0_nomal

outie_m3_m0_mal %>%
  dplyr::left_join(outie_m3_m0_nomal, by="geneset") %>%
  filter(!is.na(annotation) & !str_detect(annotation,"TBA")) ->
  outie_m3_m0


## select only significant genesets
allsigs <- unique(c(outie_m3_m0$geneset[outie_m3_m0$FDR_mal_r <= fdr_cut],
                    outie_m3_m0$geneset[outie_m3_m0$FDR_mal_c <= fdr_cut],
                    outie_m3_m0$geneset[outie_m3_m0$FDR_nomal_r <= fdr_cut],
                    outie_m3_m0$geneset[outie_m3_m0$FDR_nomal_c <= fdr_cut]))

outie_m3_m0%>%
   filter(geneset %in% allsigs) ->
outie



## set all non-significant genesets PValues to 1
outie$log10_Pval_mal_r[outie$FDR_mal_r >= fdr_cut] <- 0
outie$log10_Pval_nomal_r[outie$FDR_nomal_r >= fdr_cut] <- 0
outie$log10_Pval_mal_c[outie$FDR_mal_c >= fdr_cut] <- 0
outie$log10_Pval_nomal_c[outie$FDR_nomal_c >= fdr_cut] <- 0

outie %>%
  dplyr::arrange(annotation, log10_Pval_mal_r) %>%
  dplyr::select(c(geneset, annotation, 
                  log10_Pval_mal_r, 
                  log10_Pval_nomal_r,
                  log10_Pval_mal_c, 
                  log10_Pval_nomal_c)) %>%
  column_to_rownames("geneset") %>%
  dplyr::rename(mal_rtss=log10_Pval_mal_r) %>%
  dplyr::rename(mal_comp=log10_Pval_mal_c) %>%
  dplyr::rename(nomal_rtss=log10_Pval_nomal_r) %>%
  dplyr::rename(nomal_comp=log10_Pval_nomal_c) %>%
  as.data.frame() ->
  heat_mat

## colors for annotation
annos <- unique(heat_mat$annotation)
anno_col <- rainbow(length(annos))
anno_cols <- anno_col[as.factor(heat_mat$annotation)]
##heat_mat$annotation <- NULL


if(FALSE) {
if(DO_PDF) {pdf(file=here("output/figures/manuscript/heat_disease_M3-M0.pdf"))}

heatmap3(heat_mat, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, 
         cexRow=0.50, cexCol=1.25, showRowDendro=FALSE, margins=c(7,15),
         col=colorRampPalette(c("navy", "white", "firebrick3"))(1024),
         RowSideColors=anno_cols, RowSideLabs="")
 legend(x="bottomleft", inset=-0.11, legend=annos, fill=anno_col, bty="n", cex=0.7)
 
if(DO_PDF) {dev.off()}
}
```
  
#Down-selected disease analysis

```{r down_load}

## read in vaccine comparison results
m3_vac <- read_csv(paste0(here("output/"), "vaccine_both_dmso_M3.csv"))
m3_m0_vac <- read_csv(paste0(here("output/"), "M3-M0_both.csv"))

```
  
   
##M3 DMSO disease comparison on down-selected vaccine gene-sets  
  
```{r}

## get disease comparison genesets based on same vaccine comparison
m3_vac %>%
  filter(FDR <= fdr_cut) %$%
  geneset ->
all_sigs_ds_m3

outie_m3 %>%
  filter(geneset %in% all_sigs_ds_m3) %>%
  mutate(FDR_rtss = p.adjust(PValue_rtss, method="BH")) %>%
  mutate(FDR_comp = p.adjust(PValue_comp, method="BH")) %>%
   filter(!is.na(annotation) & !str_detect(annotation,"TBA")) %>%
  filter(FDR_rtss <= fdr_cut | FDR_comp <= fdr_cut) ->
  outie_m3_ds

write_csv(outie_m3_ds, paste0(here("output/"), "Fig3A_M3_case_control.csv"))
  
## set all non-significant genesets PValues to 1
outie_m3_ds$log10_Pval_rtss[outie_m3_ds$FDR_rtss >= fdr_cut] <- 0
outie_m3_ds$log10_Pval_comp[outie_m3_ds$FDR_comp >= fdr_cut] <- 0

outie_m3_ds %>%
  dplyr::arrange(annotation, log10_Pval_rtss, log10_Pval_comp) %>%
  dplyr::select(c(geneset, annotation, 
                  log10_Pval_rtss, 
                  log10_Pval_comp)) %>%
  column_to_rownames("geneset") %>%
  dplyr::rename(rtss=log10_Pval_rtss) %>%
  dplyr::rename(comp=log10_Pval_comp) %>%
  as.data.frame() ->
  heat_mat

## colors for annotation
annos <- unique(heat_mat$annotation)
anno_col <- rainbow(length(annos))
anno_cols <- anno_col[as.factor(heat_mat$annotation)]
heat_mat$annotation <- NULL

## only 1 significant gene-set so no heat map
if(FALSE) {
if(DO_PDF) {pdf(file=here("output/figures/manuscript/heat_disease_downselect_dmso_M3.pdf"))}

heatmap3(heat_mat, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, 
         cexRow=1, cexCol=1.25, showRowDendro=FALSE, margins=c(3,25),
         col=colorRampPalette(c("navy", "white", "firebrick3"))(1024),
         RowSideColors=anno_cols, RowSideLabs="")
 legend(x="bottomleft", inset=-0.11, legend=annos, fill=anno_col, bty="n", cex=0.7)
 
if(DO_PDF) {dev.off()}
}
```


###Driver genes from M3 disease comparison significant gene-sets (Section2)  
  
Extract driver genes from down-selected monocyte related significant gene-sets. 
 
```{r read_child}
## read in child file for rendering
child1 <- here::here("analysis/children", "gsea_plot_genes.Rmd")
childy <- read_file(child1)
```

  
```{r driver}
ds_mono_sigs <- grep("monocyte", rownames(heat_mat), value=TRUE)
p_cut <- 0.1
all_sig_genes <- NULL
```


```{r M11}
set <- ds_mono_sigs[[1]]

```
  
`r paste(knit_child(text = childy), collapse = '\n')`
</br>
   
  

```{r M81}
set <- ds_mono_sigs[[2]]

```
  
`r paste(knit_child(text = childy), collapse = '\n')`
</br>
     

##M3-M0 DMSO disease comparison on down-selected vaccine gene-sets  
  
```{r}

## get disease comparison genesets based on same vaccine comparison
m3_m0_vac %>%
  filter(FDR <= fdr_cut) %$%
  geneset ->
all_sigs_ds_m3_m0

outie_m3_m0 %>%
  filter(geneset %in% all_sigs_ds_m3) %>%
  mutate(FDR_mal_r = p.adjust(PValue_mal_r, method="BH")) %>%
  mutate(FDR_nomal_r = p.adjust(PValue_nomal_r, method="BH")) %>%
  mutate(FDR_mal_c = p.adjust(PValue_mal_c, method="BH")) %>%
  mutate(FDR_nomal_c = p.adjust(PValue_nomal_c, method="BH")) %>%
   filter(!is.na(annotation) & !str_detect(annotation,"TBA")) %>%
  filter(FDR_mal_r <= fdr_cut | FDR_nomal_r <= fdr_cut |
         FDR_mal_c <= fdr_cut | FDR_nomal_c <= fdr_cut ) ->
  outie_m3_m0_ds

write_csv(outie_m3_m0_ds, paste0(here("output/"), "Fig3B_M3-M0_rtss.csv"))
  
## set all non-significant genesets PValues to 1
outie_m3_m0_ds$log10_Pval_mal_r[outie_m3_m0_ds$FDR_mal_r >= fdr_cut] <- 0
outie_m3_m0_ds$log10_Pval_nomal_r[outie_m3_m0_ds$FDR_nomal_r >= fdr_cut] <- 0
outie_m3_m0_ds$log10_Pval_mal_c[outie_m3_m0_ds$FDR_mal_c >= fdr_cut] <- 0
outie_m3_m0_ds$log10_Pval_nomal_c[outie_m3_m0_ds$FDR_nomal_c >= fdr_cut] <- 0

outie_m3_m0_ds %>%
  dplyr::arrange(annotation, log10_Pval_mal_r, log10_Pval_nomal_r, 
                  log10_Pval_mal_c, log10_Pval_nomal_c) %>%
  dplyr::select(c(geneset, annotation, 
                  log10_Pval_mal_r, 
                  log10_Pval_nomal_r,
                  log10_Pval_mal_c, 
                  log10_Pval_nomal_c)) %>%
  column_to_rownames("geneset") %>%
  dplyr::rename(mal_rtss=log10_Pval_mal_r) %>%
  dplyr::rename(no_mal_rtss=log10_Pval_nomal_r) %>%
  dplyr::rename(mal_comp=log10_Pval_mal_c) %>%
  dplyr::rename(no_mal_comp=log10_Pval_nomal_c) %>%
  as.data.frame() ->
  heat_mat

## colors for annotation
annos <- unique(heat_mat$annotation)
anno_col <- rainbow(length(annos))
anno_cols <- anno_col[as.factor(heat_mat$annotation)]
heat_mat$annotation <- NULL

## only 1 significant gene-set so no heat map
if(FALSE) {
if(DO_PDF) {pdf(file=here("output/figures/manuscript/heat_disease_downselect_M3-M0.pdf"))}

heatmap3(heat_mat, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, 
         cexRow=1, cexCol=1.25, showRowDendro=FALSE, margins=c(1,25),
         col=colorRampPalette(c("navy", "white", "firebrick3"))(1024),
         RowSideColors=anno_cols, RowSideLabs="")
 legend(x="bottomleft", inset=-0.11, legend=annos, fill=anno_col, bty="n", cex=0.7)
 
if(DO_PDF) {dev.off()}
}
```

##All significant genes  
  
```{r}

##Bubble map  
  
```{r bubble, eval=FALSE}

## remove TBAs
 cam_visit_all %>%
   filter(!str_detect(annotation, "TBA")) ->
   cam_M

 ## reformat for bubble graph
 as.data.frame(unlist(t(table(cam_M[cam_M$FDR_both <= fdr_cut,
                                    c("Direction_both", 
                                      "annotation")])))) %>%
   dplyr::rename(Direction=Direction_both) %>%
   mutate(Dir_name=paste("both", Direction)) ->
   bub_both
   
 as.data.frame(unlist(t(table(cam_M[cam_M$FDR_rtss <= fdr_cut,
                                    c("Direction_rtss",
                                      "annotation")])))) %>%
   dplyr::rename(Direction=Direction_rtss) %>%
   mutate(Dir_name=paste("rtss", Direction)) ->
   bub_rtss
 
 as.data.frame(unlist(t(table(cam_M[cam_M$FDR_comp <= fdr_cut,
                                    c("Direction_comp",
                                      "annotation")])))) %>%
   dplyr::rename(Direction=Direction_comp) %>%
   mutate(Dir_name=paste("comp", Direction)) ->
   bub_comp
 
 as.data.frame(unlist(t(table(outie_vac[,c("Direction", "annotation")])))) %>%
   mutate(Dir_name=paste("M3", Direction)) ->
   bub_vac
 
 bub_all <- rbind(bub_vac, bub_both, bub_rtss, bub_comp)
 bub_all$Freq[bub_all$Freq == 0] <- NA
 bub_all$annotation <- factor(bub_all$annotation, levels=sort(unique(bub_all$annotation)))
 
 bub_all$Dir_name <- factor(bub_all$Dir_name, levels=c("M3 Up", "M3 Down", "both Up", "both Down", "rtss Up", "rtss Down", "comp Up", "comp Down"))
 bub_all$Direction <- relevel(bub_all$Direction, "Up")
 
 ## remove comp columns
 bub_all %>%
   filter(Dir_name == "rtss Up" | Dir_name == "rtss Down" |
           Dir_name == "M3 Up" | Dir_name == "M3 Down") ->
   bub1
 
 if(DO_PDF) {pdf(file=here("output/figures/manuscript/Fig2_bubble.pdf"))}
 ggplot(bub1, aes(x=Dir_name, y=factor(annotation, levels=rev(sort(unique(annotation)))), size=Freq,
                     color=Direction)) +
   geom_point() + xlab("") + ylab("annotation") +
   ggtitle("vaccine effect") +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   scale_color_manual(values=c("firebrick3", "navy")) 

 if(DO_PDF) {dev.off()}
 
```




