---
title: 'Manuscript: M3 DMSO'
author: "Carl Murie"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: html_document
---

```{r overall-knitr-options, echo=FALSE, results='hide'}
library(knitr)
opts_chunk$set(cache=FALSE, eval=TRUE, echo=FALSE, message=FALSE, warning=FALSE, results='asis')
```

```{r packages}
library(readr)
library(RNASeqUtilities)
library(Biobase)
library(magrittr)
library(here)
library(heatmap3)
library(DT)
library(GSEABase)
library(limma)
library(tidyverse)
library(data.table)
library(RColorBrewer)
library(heatmap3)

source(here("code", "mal_utilities.R"))
```

```{r load}
## load mal067 data
library(mal067data)
meta_dt <- as_tibble(pData(mal067_eset))

DO_PDF <- FALSE

stim="dmso"
time="M3"
```

```{r genesets}
## set up GSEA analysis
min_gene_set <- 5
btm_gtm <- getGmt(here("data/BTM_for_GSEA_20131008.gmt"))
btm_ind <- ids2indices(geneIds(btm_gtm), rownames(mal067_eset))
btm_ind <- btm_ind[sapply(btm_ind, length) > min_gene_set]

## read in BTM module annotations
btm_anno <- read_csv(here("data/btm_annotation_table_LNC.csv"))
btm_anno$full_title <- paste0(btm_anno$`Module title`, " (", btm_anno$ID, ")")

```

```{r set-parameters}

## parameters set by parent: stim, time

fdr_cut <- 0.2
p_cut <- 0.01
form_main_vac_both <- "~plate + total_reads + age + vaccine"
coeff <- "vaccinertss"
```

 
# GSEA `r toupper(stim)`: vaccine main effects 
  
**vaccine linear model:** `r form_main_vac_both`   
**coefficient:** `r coeff`  
**vaccine filtering   :**  `r time`  
  
```{r vaccine_data}
################### vaccine and disease  analysis ########################

## vaccine: select subset of expressionSet
meta_dt %>% 
  filter(stimulation == stim) %>% 
  filter(visit == time) %$%
  col_id ->
  sample_vac


## generate subset data and drop extra stimulation levels
mal_vac <- mal067_eset[, sample_vac] 
mal_vac$stimulation <- fct_drop(mal_vac$stimulation)
```


```{r sample_size}

## get sample sizes
countSampleSizes(pData(mal_vac), "stimulation", 
                 c("vaccine", "case", "age")) %>%
headerKable(labels=c("vaccine", "case", "age"), levels=c(2,3,2), 
            title="vaccine sample size")

```
  
##vaccine M3 (RTS,S - comparator)

```{r vac_analysis}
################## vaccine analysis ########################

cam_vac <- runGSEA(mal_vac, 
                   form_main_vac_both,
                   coef="vaccinertss")
numSig <- sum(cam_vac$FDR <= fdr_cut)

suffix <- paste(stim, time, sep="_")
write_csv(cam_vac, paste0(here("output/"), "vaccine_both_", suffix, ".csv"))

write_csv(cam_vac[cam_vac$FDR <= fdr_cut,], paste0(here("output/"), "Fig2A_vaccine_both_", suffix, ".csv"))


## append annotation to signficant gene-sets
cam_vac %>%
  filter(cam_vac$FDR <= fdr_cut) %>%
  left_join(btm_anno[,c("annotation", "full_title")], by=c("geneset"="full_title")) ->
  vac_sig

vac_sig %>%
  filter(!is.na(annotation)) %>%
  filter(!str_detect(annotation, "TBA")) %>%
  mutate(sign_vac=as.numeric(plyr::mapvalues(Direction, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(log10_Pval_vac=sign_vac*-log10(PValue))  ->
  outie_vac

write_csv(outie_vac, paste0(here("output/"), "M3_dmso_vaccine.csv"))
```


```{r heatme_vac}

## set all non-significant genesets PValues to 1
##outie_vac$PValue[outie_vac$FDR >= fdr_cut] <- 1


outie_vac %>%
  dplyr::arrange(annotation, log10_Pval_vac) %>%
  dplyr::select(c(geneset, annotation, log10_Pval_vac)) %>%
  column_to_rownames("geneset") %>%
  plyr::rename(c(log10_Pval_vac="vaccine")) %>%
  as.data.frame() ->
  heat_mat
heat_vac <- heat_mat

heat_mod <- cbind(heat_mat, heat_mat$vaccine)[,-1]
heat_mod$annotation <- NULL
heat_mod <- cbind(heat_mat, heat_mat[,2])[,-1]

## colors for annotation
annos <- unique(heat_mat$annotation)
anno_col <- rainbow(length(annos))
anno_cols <- anno_col[as.factor(heat_mat$annotation)]

if(DO_PDF) {pdf(file=here("output/figures/manuscript/heat_dmso_M3.pdf"))}

heatmap3(heat_mod, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, 
         cexRow=0.50, cexCol=1.25, showRowDendro=FALSE, margins=c(1,9),
         col=colorRampPalette(c("navy", "white", "firebrick3"))(1024),
         RowSideColors=anno_cols, labCol=FALSE, RowSideLabs="")
 legend(x="bottomleft", inset=-0.11, legend=annos, fill=anno_col, bty="n", cex=0.7)
 
if(DO_PDF) {dev.off()}
 
```


##vaccine M3-MO  {.tabset .tabset-fade .tabset-pills}
  
Only Bagamoya samples with the dmso stimulation are used. A consequence of this is that only subjects from the old age group are included.  Additionally only paired samples by subject are included.
  
```{r m3-m0}

form_visit_main <- "~plate + total_reads + pid + visit"
coeff <- "visitM3"

mal067_voom$targets %>%
  filter(site=="BAGAMOYO") %>%
  filter(stimulation=="dmso") %>%
  mutate(stimulation=droplevels(stimulation)) %$%
  col_id ->
  bag_ind

v_bag <- mal067_voom[,bag_ind]
v_bag$targets$stimulation <- fct_drop(v_bag$targets$stimulation)
v_bag$targets$visit <- fct_relevel(v_bag$targets$visit, "M0")


## get paired intersection between time points
pids <- intersect(v_bag$targets$pid[v_bag$targets$visit=="M0"],
                  v_bag$targets$pid[v_bag$targets$visit=="M3"])
v_bag$targets %>%
  filter(pid %in% pids) %$%
  col_id ->
  pid_ids
  
v_bag <- v_bag[,pid_ids]

v_bag$targets %>%
  countSampleSizes("stimulation", c("vaccine", "malaria_status",  "age", "visit", "case")) %>%
  headerKable(labels=c("vaccine", "disease", "age", "visit", "case"),
              levels=c(2,2,2,2,3))
```

formula: `r form_visit_main`  
coefficient: `r coeff`  
  
###rts,s and comparator together  
  
```{r visit_both}
## interaction
des <- model.matrix(formula(form_visit_main), v_bag$targets)

## run camera
cam_visit_both <- as_tibble(camera(v_bag$E, btm_ind, design=des,
                              contrast=coeff),
                       rownames="geneset")
v_both_sigs <- sum(cam_visit_both$FDR <= fdr_cut)

cam_visit_both %>%
  filter(FDR <= fdr_cut) %>%
  headerKable()

write_csv(cam_visit_both, paste0(here("output/"), "M3-M0_both.csv"))

```

###rts,s  
  
```{r visit_rtss}
## select only rts,s subjects
indy <- v_bag$targets$vaccine=="rtss"

des <- model.matrix(formula(form_visit_main), v_bag$targets[indy,])

## run camera
cam_visit_rtss <- as_tibble(camera(v_bag$E[,indy], btm_ind, design=des,
                              contrast=coeff),
                       rownames="geneset")
v_rtss_sigs <- sum(cam_visit_rtss$FDR <= fdr_cut)

cam_visit_rtss %>%
  filter(FDR <= fdr_cut) %>%
  headerKable()

write_csv(cam_visit_rtss, paste0(here("output/"), "M3-M0_rtss.csv"))

## write significant gene sets
cam_visit_rtss %>%
  dplyr::filter(FDR <= fdr_cut) %>%
  left_join(btm_anno[,c("annotation", "full_title")], by=c("geneset"="full_title")) ->
  outie

write_csv(outie, paste0(here("output/"), "Fig2B_M3-M0_rtss.csv"))

```


###comparator  
  
```{r visit_comparator}
## select only rts,s subjects
indy <- v_bag$targets$vaccine=="comparator"

des <- model.matrix(formula(form_visit_main), v_bag$targets[indy,])

## run camera
cam_visit_comp <- as_tibble(camera(v_bag$E[,indy], btm_ind, design=des,
                              contrast=coeff),
                       rownames="geneset")
v_comp_sigs <- sum(cam_visit_comp$FDR <= fdr_cut)

cam_visit_comp %>%
  filter(FDR <= fdr_cut) %>%
  headerKable()

write_csv(cam_visit_comp, paste0(here("output/"), "M3-M0_comp.csv"))
```

```{r write}

cam_visit_both %>% 
  dplyr::left_join(cam_visit_rtss[,-2], by="geneset") %>%
  dplyr::left_join(cam_visit_comp[,-2], by="geneset") %>%
  dplyr::left_join(btm_anno[,c("annotation", "full_title")],
                   by=c("geneset"="full_title"))  ->
  cam_visit_all

col_names <- c("geneset", "NGenes",
               paste0(colnames(cam_visit_both)[3:5], "_both"),
               paste0(colnames(cam_visit_both)[3:5], "_rtss"),
               paste0(colnames(cam_visit_both)[3:5], "_comp"),
               "annotation")
colnames(cam_visit_all) <- col_names
write_csv(cam_visit_all, paste0(here("output/"),
                                 "M3-M0_by_vaccine.csv"))

```

##Heatmap  
  
```{r heatme_M3-M0}

## set all non-significant genesets PValues to 1
cam_visit_all$PValue_both[cam_visit_all$FDR_both >= fdr_cut] <- 1
cam_visit_all$PValue_rtss[cam_visit_all$FDR_rtss >= fdr_cut] <- 1
cam_visit_all$PValue_comp[cam_visit_all$FDR_comp >= fdr_cut] <- 1
cam_vac$PValue[cam_vac$FDR >= fdr_cut] <- 1

cam_visit_all %>%
  dplyr::select(-c(NGenes)) %>%
  dplyr::left_join(cam_vac, by="geneset") ->
cam_total

cam_total %>%
  filter(!is.na(annotation)) %>%
  filter(!str_detect(annotation, "TBA")) %>%
  mutate(sign_vac_both=as.numeric(plyr::mapvalues(Direction_both, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(log10_Pval_both=sign_vac_both*-log10(PValue_both)) %>%
  mutate(sign_vac_rtss=as.numeric(plyr::mapvalues(Direction_rtss, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(log10_Pval_rtss=sign_vac_rtss*-log10(PValue_rtss)) %>%
  mutate(sign_vac_comp=as.numeric(plyr::mapvalues(Direction_comp, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(log10_Pval_comp=sign_vac_comp*-log10(PValue_comp)) %>%
  mutate(sign_m3=as.numeric(plyr::mapvalues(Direction, 
                                             from=c("Up", "Down"), 
                                              to=c(1,-1)))) %>%
  mutate(log10_Pval_m3=sign_m3*-log10(PValue)) ->
  outie


outie %>%
   filter(FDR_both <= fdr_cut | 
           FDR_rtss <= fdr_cut | 
           FDR_comp <= fdr_cut) %>%
  dplyr::arrange(annotation, log10_Pval_both) %>%
  dplyr::select(c(geneset, annotation, log10_Pval_both,
                  log10_Pval_rtss, log10_Pval_comp)) %>%
  column_to_rownames("geneset") %>%
  plyr::rename(c(log10_Pval_both="both")) %>%
  plyr::rename(c(log10_Pval_rtss="rtss")) %>%
  plyr::rename(c(log10_Pval_comp="comp")) %>%
  as.data.frame() ->
  heat_mat

## colors for annotation
annos <- unique(heat_mat$annotation)
anno_col <- rainbow(length(annos))
anno_cols <- anno_col[as.factor(heat_mat$annotation)]

heat_mat$annotation <- NULL

if(DO_PDF) {pdf(file=here("output/figures/manuscript/heat_M3-M0.pdf"))}

heatmap3(heat_mat, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, 
         cexRow=0.40, cexCol=1.25, showRowDendro=FALSE, margins=c(2,20),
         col=colorRampPalette(c("navy", "white", "firebrick3"))(1024),
         RowSideColors=anno_cols, RowSideLabs="", na.rm=FALSE)
 legend(x="bottomleft", inset=-0.1, legend=annos, fill=anno_col, bty="n", cex=0.7)
 
if(DO_PDF) {dev.off()}
 
 
 ## heatmap of selected annotation group
 outie %>%
   filter(annotation == "Monocytes" |
            annotation == "B Cells" |
            annotation == "T Cells" |
            annotation == "Dendritic Cells" |
            annotation == "Inflammatory/TLR/Chemokines" |
            annotation == "Mitochondria" |
            annotation == "Antiviral/IFN" |
            annotation == "NK Cells") %>%
   filter(FDR <= fdr_cut | FDR_rtss <= fdr_cut) %>%
   arrange(annotation, log10_Pval_m3) %>%
    dplyr::select(c(geneset, annotation, log10_Pval_m3,
                  log10_Pval_rtss)) %>%
  column_to_rownames("geneset") %>%
  plyr::rename(c(log10_Pval_m3="M3")) %>%
  plyr::rename(c(log10_Pval_rtss="rtss")) %>%
  as.data.frame() ->
  sel_mat
   
 ## colors for annotation
annos <- unique(sel_mat$annotation)
anno_col <- rainbow(length(annos))
anno_cols <- anno_col[as.factor(sel_mat$annotation)]

sel_mat$annotation <- NULL

if(DO_PDF) {pdf(file=here("output/figures/manuscript/Fig2_heat_select.pdf"))}

heatmap3(sel_mat, Rowv=NA, Colv=NA, scale="none", balanceColor=TRUE, 
         cexRow=1, cexCol=1.25, showRowDendro=FALSE, margins=c(2,20),
         col=colorRampPalette(c("navy", "white", "firebrick3"))(1024),
         RowSideColors=anno_cols, RowSideLabs="", na.rm=FALSE)
 legend(x="bottomleft", inset=-0.1, legend=annos, fill=anno_col, bty="n", cex=1)
 
if(DO_PDF) {dev.off()}
 
```
  
##Bubble map  
  
```{r bubble}

## remove TBAs
 cam_visit_all %>%
   filter(!str_detect(annotation, "TBA")) ->
   cam_M

 ## reformat for bubble graph
 as.data.frame(unlist(t(table(cam_M[cam_M$FDR_both <= fdr_cut,
                                    c("Direction_both", 
                                      "annotation")])))) %>%
   dplyr::rename(Direction=Direction_both) %>%
   mutate(Dir_name=paste("both", Direction)) ->
   bub_both
   
 as.data.frame(unlist(t(table(cam_M[cam_M$FDR_rtss <= fdr_cut,
                                    c("Direction_rtss",
                                      "annotation")])))) %>%
   dplyr::rename(Direction=Direction_rtss) %>%
   mutate(Dir_name=paste("rtss", Direction)) ->
   bub_rtss
 
 as.data.frame(unlist(t(table(cam_M[cam_M$FDR_comp <= fdr_cut,
                                    c("Direction_comp",
                                      "annotation")])))) %>%
   dplyr::rename(Direction=Direction_comp) %>%
   mutate(Dir_name=paste("comp", Direction)) ->
   bub_comp
 
 as.data.frame(unlist(t(table(outie_vac[,c("Direction", "annotation")])))) %>%
   mutate(Dir_name=paste("M3", Direction)) ->
   bub_vac
 
 bub_all <- rbind(bub_vac, bub_both, bub_rtss, bub_comp)
 bub_all$Freq[bub_all$Freq == 0] <- NA
 bub_all$annotation <- factor(bub_all$annotation, levels=sort(unique(bub_all$annotation)))
 
 bub_all$Dir_name <- factor(bub_all$Dir_name, levels=c("M3 Up", "M3 Down", "both Up", "both Down", "rtss Up", "rtss Down", "comp Up", "comp Down"))
 bub_all$Direction <- relevel(bub_all$Direction, "Up")
 
 ## remove comp columns
 bub_all %>%
   filter(Dir_name == "rtss Up" | Dir_name == "rtss Down" |
           Dir_name == "M3 Up" | Dir_name == "M3 Down") ->
   bub1
 
 bub1$annotation <- factor(bub1$annotation, levels=rev(sort(unique(bub1$annotation))), ordered=TRUE)
 
 bub1 %<>%
   dplyr::arrange(annotation)
 
 if(DO_PDF) {pdf(file=here("output/figures/manuscript/Fig2_bubble.pdf"))}
 ggplot(bub1, aes(x=Dir_name, y=factor(annotation, levels=rev(sort(unique(annotation)))), size=Freq,
                     color=Direction)) +
   geom_point() + xlab("") + ylab("annotation") +
   ggtitle("vaccine effect") +
   theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
   scale_color_manual(values=c("firebrick3", "navy")) 

 if(DO_PDF) {dev.off()}
 
```



